---
title: "Describing and Visualizing Continuous Variables"
execute:
  echo: true
  warning: false
  message: false
  error: true
format:
  revealjs:
    theme: simple
    css: style_lecture.css
    include-after-body: headers.html
    self-contained: true
    slide-number: true
    width: 1600
    height: 900
    df-print: kable
    html-math-method: katex
    code-fold: false
    code-tools: false
    code-annotations: hover
    incremental: false
editor: source
---

## Introduction

```{r}
#| echo: false
uwf_blue <- "#004C97"
uwf_green <- "#007A33"
uwf_marigold <- "#FFB81C"
uwf_midnight <- "#003865"
```

- Before we can "analyze data" we must know what the data *looks like*. 

- This lecture covers the first step in any analysis: **describing** a variable before you test a claim about it.

    - We will focus only on continuous variables in this lecture. 
    
    - The next lecture covers discrete variables.
    
- Ultimately, we should look first and ask (...or answer) questions later.    

## `R`? Oh No!

- This lecture begins our journey in `R`.

    - Installing `R` on your own computer:
    
        - Download `R` from CRAN: [https://cran.r-project.org/](https://cran.r-project.org/)
        
        - Download RStudio or Positron from Posit: [https://posit.co/downloads](https://posit.co/downloads)
        
    - Using the HMCSE-maintained RStudio Workbench: [https://rstudio.hmcse.uwf.edu](https://rstudio.hmcse.uwf.edu)
    
        - Only recommended for those that cannot download and install programs to their computers.
        
- There are other languages (or *tools*): python, SAS, Stata, SPSS, ...

        
## `R` Setup        

- To follow along with lecture, please install the `tidyverse` and `ssstats` packages:

```{r}
#| eval: false
install.packages("tidyverse") # <1>
if (!requireNamespace("pak", quietly = TRUE)) { # <2>
  install.packages("pak") # <3>
}
pak::pak("sealslab/ssstats") # <4>
```
1. Installs the `tidyverse` collection of packages. This only needs to run once *per computer* and *never* in a .qmd file.
2. Checks whether the `pak` package is already installed.
3. Installs `pak`, a package manager that can install packages directly from GitHub repositories.
4. Uses `pak` to install `ssstats` from the `sealslab` GitHub repository.

- Then, we must call the packages to access their functions:

```{r}
library(tidyverse) # <1>
library(ssstats) # <2>
```
1. Loads the `tidyverse` collection of packages into the current R session.
2. Loads the `ssstats` course package into the current R session.

## Data from Zootropolis

- It's your first week as a city analyst for the Zootropolis Police Department (ZPD). 

    - Deputy Chief Okonkwo walks you over to a desk, gives you a thumb drive full of data, and says: *"Your first task is to get to know this data. Tell me what's normal. Tell me what's not."*

- In particular, we will focus on the following variables:

- *academy_exam_score*: how new recruits performed on their academy entrance exam
- *hustle_earnings*: side income recruits reported earning around the district
- *district_temp_f*: average temperature logged across ZPD's patrol districts

## Data from Zootropolis

- After getting your computer set up, you read the data in:

```{r}
zootopia <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/1-zootopia.csv")
```

```{r}
#| echo: false

head(zootopia, n = 5)
```

## Describing the Center of the Data

::: {.notebook-entry}
::: {.notebook-tab}
Measures of Centrality
:::
For a set of numeric values, the number that describes the typical or central value is called a **measure of centrality**.
:::

- Note that the "center" is not a fixed idea, but a concept. We will choose from:

    - Mean
    
    - Median
    
    - Mode

## Mean

::: {.notebook-entry}
::: {.notebook-tab}
Mean
:::
The **mean** can be considered the "balance point" of the data.
:::

$$
\text{mean} = \frac{\text{sum of all values}}{\text{number of values}}
$$

- Every single value gets a vote when we calculate the mean, even the extreme ones. 

    - Extreme values will *skew* the mean, implying the mean is not robust to outliers.

## Median

::: {.notebook-entry}
::: {.notebook-tab}
Median
:::
The **median** is the middle value of the sorted data. Half the values fall below it, the other half fall above.
:::

- There is not quite a formula for the median, but the process is simple:

    - Odd number of values → the median is the literal middle one
    - Even number of values → the median is the average of the two middle values

- The median is appropriate for skewed data because it is robust to outliers.

    - We are using the *location* to determine the median -- it does not depend on the actual values of the data.
    
## Mode

::: {.notebook-entry}
::: {.notebook-tab}
Mode
:::
The **mode** is the most frequently occurring value.
:::

- A distribution can have one mode, two modes, or many modes.

    - Unimodal: just one mode.
    
    - Bimodal: two modes.
    
    - Multimodal: more than one mode.

- The mode is not reported often for continuous data, but we should be aware.

    - When looking at graphs, we may see multiple modes.
    
    - This is suggesting there may be underlying clustering or grouping.
    
## Describing the Spread of the Data

::: {.notebook-entry}
::: {.notebook-tab}
Measures of Spread
:::
For a set of numeric values, the number that describes how spread out or *dispersed* the values are is called a **measure of spread**.
:::

- We will use the variance and standard deviation to describe the spread of data.

$$
\text{variance} \approx \frac{\text{sum(deviation from the mean)}^2}{\text{number of values}}
$$

  and

$$
\text{standard deviation} =  \sqrt{\text{variance}}
$$

## Variance and Standard Deviation

$$
\text{variance} \approx \frac{\text{sum(deviation from the mean)}^2}{\text{number of values}}
$$

  and

$$
\text{standard deviation} =  \sqrt{\text{variance}}
$$

- Why do we use the standard deviation?

    - The standard deviation is in the same units as the data, while the variance is in squared units.
    
    - Suppose we are looking at the price of carrots (in Z) over time.

        - The variance would be in Z^2^ while the standard deviation would be in Z, or the same units as the data.

## Interquartile Range

- If we are using the median to describe the center of data, it is appropriate to use the interquartile range (IQR) to describe the spread of data.

$$
\text{IQR} = \text{(value at the 75th percentile)} - \text{(value at the 25th percentile)}
$$

- The IQR tells us the spread of the middle 50% of the data.

## Describing the Data (`R`)

- We will use the `mean_median()` function from the `ssstats` package to describe the center and spread of our continuous data.

```{r}
#| eval: false
dataset_name |> # <1> 
  mean_median(variable1, variable2, ...) # <2>
```
1. `dataset_name` is the name of the dataset you are working with.
2. `mean_median()` returns the mean, median, SD, and IQR for any variable(s) you list.

- We can also use `group_by()` to get the mean, median, SD, and IQR for each group of a categorical variable.

```{r}
#| eval: false
dataset_name |> # <1>
  group_by() |> # <2>
  mean_median(variable1, variable2, ...) # <3>
```
1. `dataset_name` is the name of the dataset you are working with.
2. `group_by()` allows us to group the data by categorical variables.
3. `mean_median()` returns the mean, median, SD, and IQR for any variable(s) you list.

- Note: if you need help getting started in R, please see the R module on Canvas.

## Describing the Data

- For illustrative purposes, we will describe the `academy_exam_score` variable in the `zootopia` dataset.

```{r}
zootopia |> 
  mean_median(academy_exam_score)
```

- This tells us that for academy exam scores, 

    - the mean is 77.3 with a standard deviation of 8.0, 
    - the median is 77.0 with an IQR of 10.8

## Dot Plot

::: {.notebook-entry}
::: {.notebook-tab}
Dot plot
:::
A **dot plot** places one dot for every observation along a number line. 
:::

```{r}
#| echo: false
#| fig-align: 'center'

zootopia |>
  ggplot(aes(x = academy_exam_score)) +
  geom_dotplot(fill = "#004C97", color = "white", dotsize = 0.8) +
  labs(x = "Academy Exam Score", y = NULL) +
  theme_minimal(base_size = 20) +
  theme(axis.text.y = element_blank())
```

## Dot Plot (`R`)

::: {.notebook-entry}
::: {.notebook-tab}
Dot plot
:::
A **dot plot** places one dot for every observation along a number line. 
:::

```{r}
#| eval: false
dataset_name |> # <1>
  ggplot(aes(x = outcome_name)) + # <2>
  geom_dotplot() # <3>
```
1. `dataset_name` is the name of the dataset you are working with.
2. `ggplot()` is the function that creates a plot. The `aes()` function inside it tells R which variable to plot.
3. `geom_dotplot()` is the function that creates a dot plot.

## Histogram

::: {.notebook-entry}
::: {.notebook-tab}
Histogram
:::
A **histogram** takes the same points and groups them into bins, showing how many values fall in each range.
:::

```{r}
#| echo: false
#| fig-align: 'center'
zootopia |>
  ggplot(aes(x = academy_exam_score)) +
  geom_histogram(fill = "#004C97", color = "white", bins = 10) +
  labs(x = "Academy Exam Score", y = "Count") +
  theme_minimal(base_size = 20)
```


## Histogram (`R`)

::: {.notebook-entry}
::: {.notebook-tab}
Histogram
:::
A **histogram** takes the same points and groups them into bins, showing how many values fall in each range.
:::

```{r}
#| eval: false
dataset_name |> # <1>
  ggplot(aes(x = outcome_name)) + # <2>
  geom_histogram(bins = number_of_bins) # <3>
```
1. `dataset_name` is the name of the dataset you are working with.
2. `ggplot()` is the function that creates a plot. The `aes()` function inside it tells R which variable to plot.
3. `geom_histogram()` is the function that creates a histogram.

## Density Plot

::: {.notebook-entry}
::: {.notebook-tab}
Density plot
:::
A **density plot** is a histogram with the rough edges smoothed away.
:::

```{r}
#| echo: false
#| fig-align: 'center'
zootopia |>
  ggplot(aes(x = academy_exam_score)) +
  geom_density(color = "#007A33") +
  labs(x = "Academy Exam Score", y = "Density") +
  theme_minimal()
```

## Density Plot (`R`)

::: {.notebook-entry}
::: {.notebook-tab}
Density plot
:::
A **density plot** is a histogram with the rough edges smoothed away.
:::

```{r}
#| eval: false
dataset_name |> # <1>
  ggplot(aes(x = outcome_name)) + # <2>
  geom_density() # <3>
```
1. `dataset_name` is the name of the dataset you are working with.
2. `ggplot()` is the function that creates a plot. The `aes()` function inside it tells R which variable to plot.
3. `geom_density()` is the function that creates a density plot

## Boxplot

::: {.notebook-entry}
::: {.notebook-tab}
Boxplot
:::
A **boxplot** visualizes the five number summary: minimum, Q1, median, Q3, and maximum.
:::

```{r}
#| echo: false
#| fig-align: 'center'

zootopia |>
  ggplot(aes(x = academy_exam_score)) +
  geom_boxplot(fill = uwf_marigold) +
  labs(x = "Academy Exam Score") +
  theme_minimal(base_size = 20) +
  theme(axis.text.y = element_blank())
```

## Boxplot (`R`)

::: {.notebook-entry}
::: {.notebook-tab}
Boxplot
:::
A **boxplot** visualizes the five number summary: minimum, Q1, median, Q3, and maximum.
:::

```{r}
#| eval: false

dataset_name |> # <1>
  ggplot(aes(x = outcome_name)) + # <2>
  geom_boxplot() # <3>
```
1. `dataset_name` is the name of the dataset you are working with.
2. `ggplot()` is the function that creates a plot. The `aes()` function inside it tells R which variable to plot.
3. `geom_boxplot()` is the function that creates a boxplot

## Scatterplot

::: {.notebook-entry}
::: {.notebook-tab}
Scatterplot
:::
A **scatterplot** is a graph that shows the relationship between two continuous variables.
:::

```{r}
#| echo: false
#| fig-align: 'center'

zootopia |>
  ggplot(aes(x = sleep_hours, y = academy_exam_score)) +
  geom_point(color = "#003865", size = 2) +
  labs(x = "Sleep Hours", y = "Academy Exam Score") +
  theme_minimal(base_size = 20)
```

## Scatterplot (`R`)

::: {.notebook-entry}
::: {.notebook-tab}
Scatterplot
:::
A **scatterplot** is a graph that shows the relationship between two continuous variables.
:::

```{r}
#| eval: false

dataset_name |> # <1>
  ggplot(aes(x = predictor_name, y = outcome_name)) + # <2>
  geom_point() # <3>
```
1. `dataset_name` is the name of the dataset you are working with.
2. `ggplot()` is the function that creates a plot. The `aes()` function inside it tells R which variable to plot.
3. `geom_point()` is the function that plots the points on the scatterplot.

## Example 1

- Deputy Chief Okonokwo is curious about the `hustle_earnings` data. He wants to know how much recruits are earning on the side.

- Our first task is to summarize the data.

    - Summary statistics.
    - Visualizations -- check for weird patterns.
    
        - Dot plot
        - Histogram / density
        - Boxplot

## Example 1

- Our first step in any analysis should be to literally look at the data.

    - Note: we are printing the data below for slide/lecture purposes. In "real life," we can open the dataset to examine.

```{r}
#| echo: false
zootopia |> 
  select(resident_name, department, district, hustle_earnings) |>
  head()
```

## Example 1

- Then, we look at the histogram/density plots:

```{r}
#| echo: false
#| fig-align: 'center'
zootopia |>
  ggplot(aes(x = hustle_earnings)) +
  geom_histogram(aes(y = after_stat(density)), fill = uwf_blue, color = "white", bins = 10) +
  geom_density(color = uwf_midnight, linewidth = 1) +
  labs(x = "Hustle Earnings", 
       y = "Density") +
  theme_minimal(base_size = 20)
```

## Example 1 (Code)

- Then, we look at the histogram/density plots:

```{r}
#| eval: false
zootopia |>
  ggplot(aes(x = hustle_earnings)) +
  geom_histogram(aes(y = after_stat(density)), fill = uwf_blue, color = "white", bins = 10) +
  geom_density(color = uwf_midnight, linewidth = 1) +
  labs(x = "Hustle Earnings", 
       y = "Density") +
  theme_minimal(base_size = 20)
```

## Example 1

- Now, let's look at the distribution of the data using a boxplot,

```{r}
#| echo: false
#| fig-align: 'center'
zootopia |>
  ggplot(aes(x = hustle_earnings)) +
  geom_boxplot(fill = uwf_marigold) +
  labs(x = "Hustle Earnings") +
  theme_minimal(base_size = 20) +
  theme(axis.text.y = element_blank())
```

## Example 1 (Code)

- Now, let's look at the distribution of the data using a boxplot,

```{r}
#| eval: false
zootopia |>
  ggplot(aes(x = hustle_earnings)) +
  geom_boxplot(fill = uwf_marigold) +
  labs(x = "Hustle Earnings") +
  theme_minimal(base_size = 20) +
  theme(axis.text.y = element_blank())
```

## Example 1

- We can also split the box plots by district,

```{r}
#| echo: false
#| fig-align: 'center'
zootopia |>
  ggplot(aes(x = hustle_earnings, y = district, fill = district)) +
  geom_boxplot() +
  labs(x = "Hustle Earnings", y = NULL) +
  theme_minimal(base_size = 20) +
  theme(legend.position = "none")
```

## Example 1 (Code)

- We can also split the box plots by district,

```{r}
#| eval: false
zootopia |>
  ggplot(aes(x = hustle_earnings, y = district, fill = district)) +
  geom_boxplot() +
  labs(x = "Hustle Earnings", y = NULL) +
  theme_minimal(base_size = 20) +
  theme(legend.position = "none")
```

## Example 1

- The boxplot confirms that there are extreme values.

- Exploring the extreme values,

```{r}
zootopia |> 
  select(resident_name, department, district, hustle_earnings) |>
  filter(hustle_earnings > 80)
```

## Example 1

- Now, we can find the descriptive statistics. That requires `mean_median()`,

```{r}
zootopia |> 
  mean_median(hustle_earnings)
```

- The difference between the mean and median suggests the data is skewed, as we saw with the histogram.

- The overall mean is helpful, but we should look at this more granularly to give the city better information.
        
## Example 1

- Let's now find the descriptive statistics by district. That requires `mean_median()`, but now we add in `group_by()`,

```{r}
zootopia |> 
  group_by(district) |> 
  mean_median(hustle_earnings)
```

## Visualizing the Summary Statistics

- Until now, we've focused on describing the *entire distribution* of a variable.

- Sometimes, we want to communicate just a single number (e.g., mean) and how much to trust it. 

    - We will graph a single value with a range around it showing uncertainty or spread.
    
- We have two choices to show the uncertainty:

    - Standard deviation
    
        - This is the spread of the data around the mean.
    
    - Standard error
    
        - This describes how precise the estimate of the mean is.

## Standard Error

::: {.notebook-entry}
::: {.notebook-tab}
Standard error
:::
**Standard error (SE)** describes how *precise* the estimate of the mean is
:::

$$
\text{SE} = \frac{\text{SD}}{\sqrt{n}}
$$

- This is the standard deviation adjusted for the sample size.

## Visualizing the Mean

```{r}
#| echo: false
#| fig-align: 'center'
earnings_summary <- zootopia |>
  summarise(
    mean_earnings = mean(hustle_earnings), 
    sd_earnings = sd(hustle_earnings),
    n = n(),
    se_earnings = sd_earnings / sqrt(n) 
  )

earnings_summary |>
  pivot_longer(
    cols = c(sd_earnings, se_earnings), 
    names_to = "error_type",
    values_to = "error_value"
  ) |>
  mutate(error_type = if_else(error_type == "sd_earnings", "SD", "SE")) |>
  ggplot(aes(x = error_type, y = mean_earnings, color = error_type)) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(ymin = mean_earnings - error_value, ymax = mean_earnings + error_value), 
    width = 0.1
  ) +
  scale_color_manual(values = c(uwf_blue, uwf_marigold)) +
  labs(x = NULL, y = "Hustle Earnings") +
  theme_minimal(base_size = 20) +
  theme(legend.position = "none")
```

## Visualizing the Mean

- Finally, we can of course, split this out by district,

```{r}
#| echo: false
#| fig-align: 'center'
zootopia |>
  group_by(district) |> # <1>
  summarise(
    mean_earnings = mean(hustle_earnings),
    sd_earnings = sd(hustle_earnings),
    n = n(),
    se_earnings = sd_earnings / sqrt(n)
  ) |>
  ggplot(aes(x = district, y = mean_earnings, color = district)) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(ymin = mean_earnings - se_earnings, ymax = mean_earnings + se_earnings), # <2>
    width = 0.15
  ) +
  labs(x = NULL, y = "Hustle Earnings") +
  theme_minimal(base_size = 20) +
  theme(legend.position = "none")
```

## Visualizing the Mean

- Some notes from a professional point of view:

    - Some folks like to use bar graphs to represent the mean. This is inaccurate.
    
    - How to choose between using SD and SE? I always use SE.
    
        - Preview for the next module: the SE is used in the calculation of confidence intervals.
        
## Wrap Up

- We have covered basics for describing and visualizing continuous variables.

    - Center:
    
        - Mean
        - Median
        - Mode
        
    - Spread:
    
        - Variance
        - Standard deviation
        - (Standard error)
        - Interquartile range

