---
title: "Describing and Visualizing Categorical Variables"
execute:
  echo: true
  warning: false
  message: false
  error: true
format:
  revealjs:
    theme: simple
    css: style_lecture.css
    include-after-body: headers.html
    self-contained: true
    slide-number: true
    width: 1600
    height: 900
    df-print: kable
    html-math-method: katex
    code-fold: false
    code-tools: false
    code-annotations: hover
    incremental: false
editor: source
---

## Introduction

```{r}
#| echo: false
uwf_blue <- "#004C97"
uwf_green <- "#007A33"
uwf_marigold <- "#FFB81C"
uwf_midnight <- "#003865"
```

- Last lecture, we described **continuous** variables: things measured on a number line, like a score or an income.

- Today we describe **categorical** variables: things that sort recruits into *groups* rather than measuring them on a scale.

    - There is no "average" district the way there is an average exam score.

- The goal is the same as always: look first, ask questions later.

## Data from Zootropolis

- You're back at your desk at ZPD. Deputy Chief Okonkwo drops by again: *"Numbers are great, but I also need to know who's where. Break the recruits down by group for me."*

- Today we'll focus on:

- *department*: whether a recruit is assigned to Patrol or Desk duty
- *district*: which of the four ZPD districts a recruit works in (Rainforest District, Downtown, Sahara Square, Tundratown)
- *academy_status*: whether a recruit passed or failed the academy entrance exam

## Data from Zootropolis

```{r}
library(tidyverse)
library(ssstats)
```

```{r}
zootopia <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/1-zootopia.csv")
```

```{r}
#| echo: false
zootopia |>
  select(resident_name, department, district, academy_status) |>
  head(n = 5)
```

## Describing Categorical Data

::: {.notebook-entry}
::: {.notebook-tab}
Categorical Data
:::
For categorical variables, we describe data by counting how many observations fall into each category, not by averaging values.
:::

- We have two main tools for describing a categorical variable:

    - Frequency
    - Relative frequency (proportion)

- There's no "mean" or "standard deviation" for a category like Sahara Square — it wouldn't mean anything to average two districts together.

## Frequency

::: {.notebook-entry}
::: {.notebook-tab}
Frequency
:::
The **frequency** of a category is simply a count: how many observations fall into that group.
:::

- Frequencies answer the question: *how many?*

    - How many recruits are in Patrol?
    - How many recruits work in Sahara Square?

- Frequencies are useful, but they depend on the size of your dataset — 18 Patrol recruits sounds different in a department of 30 than in a department of 3,000.

## Relative Frequency (Proportion)

::: {.notebook-entry}
::: {.notebook-tab}
Relative Frequency
:::
The **relative frequency**, or **proportion**, is the frequency divided by the total number of observations.
:::

$$
\text{proportion} = \frac{\text{count in category}}{\text{total number of observations}}
$$

- Proportions answer the question: *what share?*

    - What share of recruits are in Patrol?
    - What share of recruits work in Sahara Square?

- Because proportions are standardized, they let us compare groups of very different sizes — 60% Patrol means the same thing whether you have 30 recruits or 3,000.

## Mode, Revisited

- Recall from last lecture: the **mode** is the most frequently occurring value.

- For categorical data, the mode is the *most natural* measure of centrality — it's really the only one that makes sense.

    - The modal district is whichever district has the most recruits.
    - A department (Patrol/Desk) can only ever be unimodal or bimodal (tied), since there are only two categories.

## What About Spread?

- For continuous data, we described spread with variance, standard deviation, and the IQR.

- Categories don't have a distance between them — Sahara Square isn't "two units" away from Downtown the way 80 is two units away from 78.

    - So variance and standard deviation don't apply to categorical data.

- Instead, when we talk about "spread" for categorical data, we're really asking: *are the observations spread evenly across categories, or concentrated in just one or two?*

    - We'll answer that visually, with bar charts, rather than with a single number.

## Describing the Data (`R`)

- We will use the `n_pct()` function from the `ssstats` package to describe the frequency and proportion of our categorical data.

```{r}
#| eval: false
dataset_name |> # <1>
  n_pct(variable1, variable2, ...) # <2>
```
1. `dataset_name` is the name of the dataset you are working with.
2. `n_pct()` returns the count (n) and percentage for each category of any variable(s) you list.

- Just like `mean_median()`, we can use `group_by()` to get counts and percentages broken down by another categorical variable.

```{r}
#| eval: false
dataset_name |> # <1>
  group_by() |> # <2>
  n_pct(variable1, variable2, ...) # <3>
```
1. `dataset_name` is the name of the dataset you are working with.
2. `group_by()` allows us to group the data by a categorical variable before counting.
3. `n_pct()` returns the count and percentage for each category, within each group.

- Note: if you need help getting started in R, please see the R module on Canvas.

## Describing the Data

- For illustrative purposes, we will describe the `department` variable in the `zootopia` dataset.

```{r}
zootopia |>
  n_pct(department)
```

- This tells us that of the 30 recruits,

    - 18 (60.0%) are in Patrol,
    - 12 (40.0%) are on Desk duty.

## Bar Chart (Counts)

::: {.notebook-entry}
::: {.notebook-tab}
Bar chart
:::
A **bar chart** shows one bar per category, with the height of the bar equal to the number (or percentage) of observations in that category.
:::

```{r}
#| echo: false
#| fig-align: 'center'
zootopia |>
  ggplot(aes(x = department, fill = department)) +
  geom_bar() +
  scale_fill_manual(values = c(uwf_blue, uwf_marigold)) +
  labs(x = "Department", y = "Count") +
  theme_minimal(base_size = 20) +
  theme(legend.position = "none")
```

## Bar Chart (Counts) (`R`)

::: {.notebook-entry}
::: {.notebook-tab}
Bar chart
:::
A **bar chart** shows one bar per category, with the height of the bar equal to the number (or percentage) of observations in that category.
:::

```{r}
#| eval: false
dataset_name |> # <1>
  ggplot(aes(x = category_name)) + # <2>
  geom_bar() # <3>
```
1. `dataset_name` is the name of the dataset you are working with.
2. `ggplot()` is the function that creates a plot. The `aes()` function inside it tells R which variable to plot.
3. `geom_bar()` counts the observations in each category and draws a bar for each — you don't need to calculate the counts yourself.

## Bar Chart (Proportions)

- Sometimes we'd rather show the *share* of recruits in each category instead of the raw count.

```{r}
#| echo: false
#| fig-align: 'center'
zootopia |>
  ggplot(aes(x = department, y = after_stat(count) / sum(after_stat(count)), fill = department)) +
  geom_bar() +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c(uwf_blue, uwf_marigold)) +
  labs(x = "Department", y = "Percent of Recruits") +
  theme_minimal(base_size = 20) +
  theme(legend.position = "none")
```

## Bar Chart (Proportions) (`R`)

- Sometimes we'd rather show the *share* of recruits in each category instead of the raw count.

```{r}
#| eval: false
dataset_name |> # <1>
  ggplot(aes(x = category_name, y = after_stat(count) / sum(after_stat(count)))) + # <2>
  geom_bar() + # <3>
  scale_y_continuous(labels = scales::percent) # <4>
```
1. `dataset_name` is the name of the dataset you are working with.
2. `after_stat(count) / sum(after_stat(count))` converts the raw counts `geom_bar()` calculates into a proportion of the total.
3. `geom_bar()` still does the counting — we're just changing what's plotted on the y-axis.
4. `scale_y_continuous(labels = scales::percent)` relabels the y-axis as a percentage instead of a decimal.

## A Word on Pie Charts

- You've almost certainly seen categorical data shown as a pie chart. We are not going to use them in this course.

- The human eye is bad at comparing *angles* and *areas* — it's very good at comparing the *height* of bars lined up on the same baseline.

    - Try to tell, at a glance, whether Sahara Square's slice is bigger than Downtown's. Now try it on the bar chart from a few slides ago.

- A bar chart shows the same information as a pie chart, but it's easier to read accurately — and it scales to more categories without turning into a mess of slivers.

    - As a working analyst, if a pie chart crosses your desk, a bar chart is almost always the fix.

## Two-Way Tables

::: {.notebook-entry}
::: {.notebook-tab}
Two-way table
:::
A **two-way table** shows how two categorical variables relate to one another, by counting observations at every combination of categories.
:::

- So far we've described one categorical variable at a time.

- Just like a scatterplot let us compare two continuous variables, a two-way table lets us compare two categorical variables.

    - Is department evenly split across every district, or do some districts lean more Patrol or more Desk?

## Two-Way Tables (`R`)

- We build a two-way table the same way we built a grouped summary before — with `group_by()` and `n_pct()`.

```{r}
#| eval: false
dataset_name |> # <1>
  group_by(grouping_variable) |> # <2>
  n_pct(second_variable) # <3>
```
1. `dataset_name` is the name of the dataset you are working with.
2. `group_by()` splits the data into groups based on the first categorical variable.
3. `n_pct()` then counts and calculates percentages of the second variable, *within* each group.

## Grouped Bar Chart

::: {.notebook-entry}
::: {.notebook-tab}
Grouped bar chart
:::
A **grouped bar chart** places bars for a second category side by side within each level of the first category, so you can visually compare the split.
:::

```{r}
#| eval: false
dataset_name |> # <1>
  ggplot(aes(x = category_1, fill = category_2)) + # <2>
  geom_bar(position = "dodge") # <3>
```
1. `dataset_name` is the name of the dataset you are working with.
2. `fill = category_2` colors the bars by the second categorical variable.
3. `position = "dodge"` places the bars for each `category_2` level side by side, instead of stacking them.

## Example 1

- Deputy Chief Okonkwo wants to know how recruits are distributed across ZPD's four districts.

- Our first task is to summarize the data.

    - Summary statistics.
    - Visualizations — check for imbalance across districts.

        - Bar chart (counts)
        - Bar chart (proportions)

## Example 1

- As always, our first step is to look at the data itself.

```{r}
#| echo: false
zootopia |>
  select(resident_name, department, district) |>
  head()
```

## Example 1

- Next, the bar chart of counts:

```{r}
#| echo: false
#| fig-align: 'center'
zootopia |>
  ggplot(aes(x = district, fill = district)) +
  geom_bar() +
  scale_fill_manual(values = c(uwf_blue, uwf_green, uwf_marigold, uwf_midnight)) +
  labs(x = NULL, y = "Count") +
  theme_minimal(base_size = 20) +
  theme(legend.position = "none")
```

## Example 1 (Code)

- Next, the bar chart of counts:

```{r}
#| eval: false
zootopia |>
  ggplot(aes(x = district, fill = district)) +
  geom_bar() +
  labs(x = NULL, y = "Count") +
  theme_minimal(base_size = 20) +
  theme(legend.position = "none")
```

## Example 1

- Now the bar chart of proportions:

```{r}
#| echo: false
#| fig-align: 'center'
zootopia |>
  ggplot(aes(x = district, y = after_stat(count) / sum(after_stat(count)), fill = district)) +
  geom_bar() +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c(uwf_blue, uwf_green, uwf_marigold, uwf_midnight)) +
  labs(x = NULL, y = "Percent of Recruits") +
  theme_minimal(base_size = 20) +
  theme(legend.position = "none")
```

## Example 1 (Code)

- Now the bar chart of proportions:

```{r}
#| eval: false
zootopia |>
  ggplot(aes(x = district, y = after_stat(count) / sum(after_stat(count)), fill = district)) +
  geom_bar() +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Percent of Recruits") +
  theme_minimal(base_size = 20) +
  theme(legend.position = "none")
```

## Example 1

- Now we find the exact numbers with `n_pct()`,

```{r}
zootopia |>
  n_pct(district)
```

- Sahara Square has the most recruits (9, or 30.0%), and Tundratown has the fewest (6, or 20.0%) — but no district is dramatically over- or under-staffed compared to the others.

## Example 1

- Deputy Chief Okonkwo has a follow-up: *"Is Patrol/Desk assignment spread the same way across every district, or does that break down differently district by district?"*

- That requires a two-way table. We add `group_by(district)` before `n_pct(department)`,

```{r}
zootopia |>
  group_by(district) |>
  n_pct(department)
```

## Example 1

- Let's visualize that same breakdown with a grouped bar chart,

```{r}
#| echo: false
#| fig-align: 'center'
zootopia |>
  ggplot(aes(x = district, fill = department)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c(uwf_blue, uwf_marigold)) +
  labs(x = NULL, y = "Count") +
  theme_minimal(base_size = 20)
```

## Example 1 (Code)

- Let's visualize that same breakdown with a grouped bar chart,

```{r}
#| eval: false
zootopia |>
  ggplot(aes(x = district, fill = department)) +
  geom_bar(position = "dodge") +
  labs(x = NULL, y = "Count") +
  theme_minimal(base_size = 20)
```

## Example 1

- A few things stand out from the two-way table and grouped bar chart:

    - Tundratown is the only district where Desk (4) outnumbers Patrol (2) — every other district leans Patrol.
    - Downtown and Sahara Square both have a roughly 2-to-1 Patrol-to-Desk split.

- On its own, the overall 60/40 Patrol/Desk split (from our first `n_pct(department)`) hides this — it's only by breaking things down by district that the Tundratown pattern shows up.

    - This is the same lesson from last lecture's `hustle_earnings` example: the overall summary and the by-group summary can each tell a different part of the story. Good analysts check both.

## Wrap Up

- We have covered basics for describing and visualizing categorical variables.

    - One categorical variable:

        - Frequency (count)
        - Relative frequency (proportion)
        - Bar chart (counts)
        - Bar chart (proportions)

    - Two categorical variables:

        - Two-way table
        - Grouped bar chart

- Next up: turning these descriptions into a testable claim about a proportion.
