---
title: "Describing and Visualizing Categorical Variables"
execute:
  echo: true
  warning: false
  message: false
  error: true
format:
  revealjs:
    theme: simple
    css: style_lecture.css
    include-after-body: headers.html
    self-contained: true
    slide-number: true
    width: 1600
    height: 900
    df-print: kable
    html-math-method: katex
    code-fold: false
    code-tools: false
    code-annotations: hover
    incremental: false
editor: source
---

## Introduction

```{r}
#| echo: false
uwf_blue <- "#004C97"
uwf_green <- "#007A33"
uwf_marigold <- "#FFB81C"
uwf_midnight <- "#003865"
```

- At ZPD, Deputy Chief Okonkwo drops by again: *"Numbers are great, but I also need to know who's where. Break the recruits down by group for me."*

- In the last lecture, we described **continuous** variables: things measured on a number line, like a score or an income.

- Today we will describe **categorical** variables: things that sort recruits into *groups* rather than measuring them on a scale.

    - There is no "average" district the way there is an average exam score.

## `R` Set Up

```{r}
library(tidyverse)
library(ssstats)
zootopia <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/1-zootopia.csv")
```

```{r}
#| echo: false
zootopia |>
  select(resident_name, department, district, academy_status) |>
  head(n = 5)
```




## Describing Categorical Data

::: {.notebook-entry}
::: {.notebook-tab}
Categorical Data
:::
For **categorical variables**, we describe data by counting how many observations fall into each category, not by averaging values.
:::

- We have two main tools for describing a categorical variable:

    - Frequency
    - Relative frequency (proportion)


## Frequency

::: {.notebook-entry}
::: {.notebook-tab}
Frequency
:::
The **frequency** of a category is simply a count: how many observations fall into that group.
:::

- Frequencies answer the question: *how many?*

    - How many recruits are in Patrol?
    - How many recruits work in Sahara Square?

- Frequencies are useful, but interpreting them depends on the size of the dataset.

## Relative Frequency 

::: {.notebook-entry}
::: {.notebook-tab}
Relative Frequency
:::
The **relative frequency** (proportion), is the frequency divided by the total number of observations.
:::

$$
\text{proportion} = \frac{\text{count in category}}{\text{total number of observations}}
$$

- Proportions answer the question: *what share?*

    - What share of recruits are in Patrol?
    - What share of recruits work in Sahara Square?

- Because proportions are standardized, they allow us to compare groups of very different sizes.

## Describing the Data (`R`)

- We will use the `n_pct()` function from the `ssstats` package to describe the frequency and proportion of our categorical data.

```{r}
#| eval: false
dataset_name |> # <1>
  n_pct(variable) # <2>
```
1. `dataset_name` is the name of the dataset you are working with.
2. `n_pct()` returns the count (n) and percentage for each category of the variable you list.

- Just like `mean_median()`, we can use `group_by()` to get counts and percentages broken down by another categorical variable.

- Note: if you need help getting started in R, please see the R module on Canvas.

## Describing the Data

- For illustrative purposes, we will describe the `department` variable in the `zootopia` dataset.

```{r}
zootopia |>
  n_pct(department)
```

- This tells us that of the 30 recruits,

    - 18 (60.0%) are in Patrol,
    - 12 (40.0%) are on Desk duty.

## Bar Graphs 

::: {.notebook-entry}
::: {.notebook-tab}
Bar chart
:::
A **bar chart** shows one bar per category, with the height of the bar equal to the number (or percentage) of observations in that category.
:::

```{r}
#| echo: false
#| fig-align: 'center'
p1 <- zootopia |>
  ggplot(aes(x = department, fill = department)) +
  geom_bar() +
  scale_fill_manual(values = c(uwf_blue, uwf_marigold)) +
  labs(x = "Department", y = "Count") +
  theme_minimal(base_size = 20) +
  theme(legend.position = "none")

p2 <- zootopia |>
  ggplot(aes(x = department, y = after_stat(count) / sum(after_stat(count)), fill = department)) +
  geom_bar() +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c(uwf_blue, uwf_marigold)) +
  labs(x = "Department", y = "Percent") +
  theme_minimal(base_size = 20) +
  theme(legend.position = "none")

library(ggpubr)

ggarrange(p1, p2, ncol = 2)
```

## Bar Graphs (`R`)

::: {.notebook-entry}
::: {.notebook-tab}
Bar chart
:::
A **bar chart** shows one bar per category, with the height of the bar equal to the number (or percentage) of observations in that category.
:::

```{r}
#| eval: false
dataset_name |> # <1>
  ggplot(aes(x = category_name)) + # <2>
  geom_bar() # <3>
```
1. `dataset_name` is the name of the dataset you are working with.
2. `ggplot()` is the function that creates a plot. The `aes()` function inside it tells R which variable to plot.
3. `geom_bar()` counts the observations in each category and draws a bar for each — you don't need to calculate the counts yourself.

- Sometimes we'd rather show the *share* of recruits in each category instead of the raw count.

```{r}
#| eval: false
dataset_name |> # <1>
  ggplot(aes(x = category_name, 
             y = after_stat(count) / sum(after_stat(count)))) + # <2>
  geom_bar() + # <3>
  scale_y_continuous(labels = scales::percent) # <4>
```
1. `dataset_name` is the name of the dataset you are working with.
2. `after_stat(count) / sum(after_stat(count))` converts the raw counts into a proportion of the total.
3. `geom_bar()` creates a bar graph.
4. `scale_y_continuous(labels = scales::percent)` relabels the y-axis as a percentage instead of a decimal.

## Contingency Tables

::: {.notebook-entry}
::: {.notebook-tab}
Contingency Table
:::
A **two-way table** shows how two categorical variables relate to one another, by counting observations at every combination of categories.
:::

- So far we've described one categorical variable at a time.

- Just like a scatterplot let us compare two continuous variables, a contingency table lets us compare two categorical variables.

    - Is department evenly split across every district, or do some districts lean more Patrol or more Desk?

## Contingency Tables (`R`)

- We can construct a contingency table using `n_pct()`.

```{r}
#| eval: false
dataset_name |> # <1>
  n_pct(row_variable, column_variable) # <2>
```
1. `dataset_name` is the name of the dataset you are working with.
2. `n_pct()` constructs an R x C contingency table.

- Note that `n_pct()` automatically returns the *column* percentage.

## Example 1

- Let's describe the characteristics of the recruits in the `zootopia` dataset.

    - Department
    
    - District
    
    - Academy status
    
## Example 1 

- Constructing the frequency distribution for department,

```{r}
zootopia %>% n_pct(department)
```

## Example 1 

- Constructing the frequency distribution for district,

```{r}
zootopia %>% n_pct(district)
```

## Example 1 

- Constructing the frequency distribution for academy status,

```{r}
zootopia %>% n_pct(academy_status)
```

## Example 1

- Constructing contingency tables,

```{r}
zootopia %>% n_pct(district, department)
```

## Example 1

- Constructing contingency tables,

```{r}
zootopia %>% n_pct(district, academy_status)
```

## Example 1

- Constructing contingency tables,

```{r}
zootopia %>% n_pct(department, academy_status)
```

## Example 1

- Looking at this graphically,

```{r}
#| echo: false
#| fig-align: 'center'

p1 <- zootopia |>
  ggplot(aes(x = department, fill = academy_status)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c(uwf_blue, uwf_green)) +
  labs(x = NULL, y = "Count", fill = "Academy Status") +
  theme_minimal(base_size = 20)

p2 <- zootopia |>
  ggplot(aes(x = department, y = after_stat(count) / sum(after_stat(count)), fill = academy_status)) +
  geom_bar(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c(uwf_blue, uwf_green)) +
  labs(x = NULL, y = "Percent", fill = "Academy Status") +
  theme_minimal(base_size = 20)

ggpubr::ggarrange(p1, p2, ncol = 2, common.legend = TRUE, legend = "bottom")
```


## Wrap Up

- We have covered basics for describing and visualizing categorical variables.

    - Frequencies and proportions
        - Contingency tables
    - Bar graphs
