---
title: "One-Sample Proportions"
execute:
  echo: true
  warning: false
  message: false
  error: true
format:
  revealjs:
    theme: simple
    css: style_lecture.css
    include-after-body: headers.html
    self-contained: true
    slide-number: true
    center: false
    width: 1600
    height: 900
    df-print: kable
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
editor: source
---

## Introduction

## `R` Setup 

```{r}
library(tidyverse)
library(ssstats)
```

## Data Setup 

- Recall our Zootopia data,

```{r}
zootopia <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/1-zootopia.csv")
```

```{r}
#| echo: false

head(zootopia, n = 5)
```

## Confidence Interval for $\pi$

- **(1--$\alpha$)100% CI for a population proportion, $\pi$**

$$\hat{p} \pm z_{\alpha/2} \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$

- $\hat{p}$ is the sample proportion (the point estimate)

- $z_{\alpha/2}$ comes from the standard normal distribution (the critical value)

- $\hat{p}(1-\hat{p})$ is the variance of the Bernoulli distribution and $n$ is the sample size ($\sqrt{\hat{p}(1-\hat{p})/n}$ is the standard error of the mean)

## Hypothesis Test for $\pi$

- **Hypotheses**

::: {.hyp-columns .columns}

::: {.column width="30%"}
**Two-Tailed**

$H_0: \pi = \pi_0$

$H_1: \pi \ne \pi_0$
:::

::: {.column width="30%"}
**Left-Tailed**

$H_0: \pi \ge \pi_0$

$H_1: \pi < \pi_0$
:::

::: {.column width="30%"}
**Right-Tailed**

$H_0: \pi \le \pi_0$

$H_1: \pi > \pi_0$
:::

:::


- $\pi$ is the population proportion,

- $\pi_0$ is the hypothesized value of $\pi$.


## Hypothesis Test for $\pi$

- **Test Statistic**

$$z_0 = \frac{\hat{p} - \pi_0}{\sqrt{\frac{\pi_0(1-\pi_0)}{n}}}$$

- where

    - $\hat{p}$ is the sample proportion

    - $\pi_0$ is the hypothesized value of $\pi$

    - $n$ is the sample size

## Statistical Inference on One Proportion (`R`)

- We will use the `one_proportion()` function from `library(ssstats)` to perform statistical inference.

```{r}
#| eval: false
dataset_name %>% one_proportion(outcome = variable_name,
                                event = event_name_or_number,
                                p = hypothesized_value,
                                alternative = "two" | "less" | "greater",
                                alpha = alpha_level)
```

- where

    - `outcome` is the variable name of the outcome variable
    
    - `event` is the name or number of the event of interest
    
    - `p` is the hypothesized value of the population proportion (default = 0.05)
    
    - `alternative` is the alternative hypothesis (default = "two")

    - `alpha` is the significance level (default = 0.05)


## Example 1

- ZPD's official staffing policy states that 50% of officers are assigned to Patrol, with the rest working Desk duty. As Zootopia's data analyst, you believe thatatrol seems to make up a lot more than 50% of the department these days. We will use the `zootopia` dataset to answer:

    - Has the true proportion of officers in Patrol increased from 50%?
    
- First, we need to describe the data.

    - We start with point and interval estimates.
    
- Then, we will perform a hypothesis test to answer the question.

## Example 1
    
- Point and interval estimates:

```{r}
zootopia %>% one_proportion(outcome = department)
```

- Oops! We need to make sure we specify the *event* of interest. Let's see how it's coded in the `department` variable.

```{r}
zootopia %>% n_pct(department)
```

## Example 1
    
- Point and interval estimates:

```{r}
zootopia %>% one_proportion(outcome = department,
                            event = "Patrol")
```

- The estimated proportion of officers in Patrol is 60% (95% CI: 42.5%, 77.5%).

## Example 1

- Then, we will perform a hypothesis test to answer the question.    

    - Has the true proportion of officers in Patrol increased from 50%?
    
- How do we translate this into hypotheses?  

    - $H_0: \pi < 0.5$
    - $H_1: \pi \ge 0.5$
    
- For the `one_proportion()` function, we will set 

    - `p = 0.5`
    - `alternative = "greater"`.    
    

## Example 1
    
- Has the true proportion of officers in Patrol increased from 50%?

```{r}
zootopia %>% one_proportion(outcome = department,
                            event = "Patrol",
                            p = 0.5,
                            alternative = "greater",
                            alpha = 0.05)
```    

## Example 1

- **Hypotheses**

    - $H_0: \pi < 0.5$
    - $H_1: \pi \ge 0.5$
    
- **Test Statistic and *p*-Value**

    - $z_0 = 1.1$
    - $p = 0.137$
    
- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Fail to reject $H_0$. There is not sufficient evidence to suggest that the current proportion of officers in Patrol has increased from 50%. 
  

## Example 2

- Zootopia's Housing Authority claims that 40% of officers live in Downtown. As data analyst, you believe Downtown might actually make up a smaller proportion of the roster than the Authority's number suggests. We will use the `zootopia` dataset to answer:

    - Is the true proportion of officers living in Downtown actually less than 40%?
    
- First, we need to describe the data.

    - We start with point and interval estimates.    

- Then, we will perform a hypothesis test to answer the question.

## Example 2

- Looking at what is in `district`,

```{r}
zootopia %>% n_pct(district)
```

- We know that the event of interest is coded as `Downtown`.

## Example 2

- Point and interval estimates:

```{r}
zootopia %>% one_proportion(outcome = district,
                            event = "Downtown")
```

- The estimated proportion of officers living in Downtown is 26.7% (95% CI: 10.8%, 42.5%).

## Example 2

- Then, we will perform a hypothesis test to answer the question.    

    - Is the true proportion of officers living in Downtown actually less than 40%?
    
- How do we translate this into hypotheses?  

    - $H_0: \pi > 0.4$
    - $H_1: \pi \le 0.4$
    
- For the `one_proportion()` function, we will set 

    - `p = 0.4`
    - `alternative = "less"`.    
    

## Example 2
    
- Is the true proportion of officers living in Downtown actually less than 40%?

```{r}
zootopia %>% one_proportion(outcome = district,
                            event = "Downtown",
                            p = 0.4,
                            alternative = "less",
                            alpha = 0.05)
```    

## Example 2

- **Hypotheses**

    - $H_0: \pi < 0.4$
    - $H_1: \pi \ge 0.4$
    
- **Test Statistic and *p*-Value**

    - $z_0 = - 1.49$
    - $p = 0.068$
    
- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Fail to reject $H_0$. There is not sufficient evidence to suggest that the current proportion of officers living in Downtown is less than 40%.

## Wrap Up

