---
title: "Independent<br>Means and Medians"
execute:
  echo: true
  warning: false
  message: false
  error: true
format:
  revealjs:
    theme: simple
    css: style_lecture.css
    include-after-body: headers.html
    self-contained: true
    slide-number: true
    center: false
    width: 1600
    height: 900
    df-print: kable
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
    code-annotations: hover
editor: source
---

## Introduction
    
- In the last module, we reviewed statistical inference for one sample.
    
- In this module, we will focus on tests for two samples.
    - Two-sample independent means ($\mu_1-\mu_2$)
    - Two-sample dependent means ($\mu_d$)
    - Two-sample proportions ($\pi_1-\pi_2$)
    - Two-sample variances ($\sigma^2_1/\sigma^2_2$)
    
- In this lecture: two-sample independent means ($\mu_1-\mu_2$)  

## `R` Setup 

```{r}
library(tidyverse)
library(ssstats)
```

- Recall our Zootopia data,

```{r}
zootopia <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/1-zootopia.csv")
```

```{r}
#| echo: false

head(zootopia, n = 3)
```

## Confidence Interval for $\mu_1-\mu_2$

- (1--$\alpha$)100% CI for a population mean, $\mu_1-\mu_2$

$$\bar{y} \pm t_{\alpha/2, n-1} \frac{s}{\sqrt{n}}$$

- $t$ has $n-1$ degrees of freedom (df). 


## Hypothesis Test for $\mu_1-\mu_2$

**Hypotheses**

::: {.hyp-columns .columns}

::: {.column width="30%"}
**Two-Tailed**

$H_0: \mu_1-\mu_2 = \mu_0$

$H_1: \mu_1 - \mu_2 \ne \mu_0$
:::

::: {.column width="30%"}
**Left-Tailed**

$H_0: \mu_1-\mu_2 \ge \mu_0$

$H_1: \mu_1-\mu_2 < \mu_0$
:::

::: {.column width="30%"}
**Right-Tailed**

$H_0: \mu_1-\mu_2 \le \mu_0$

$H_1: \mu_1-\mu_2 > \mu_0$
:::

:::

- $\mu_i$ is the population mean for group $i$,

- $\mu_0$ is the hypothesized difference between the groups

## Hypothesis Test for $\mu_1-\mu_2$

**Test Statistic:**

$$
t_0 = \frac{(\bar{x}_1-\bar{x}_2)-\mu_0}{{\sqrt{\frac{s_1^2}{n_1} + \frac{s^2_2}{n_2}}}} \sim t_{\text{df}}
$$

- where
    - $\bar{x}_i$ is the mean for group $i$
    - $\mu_0$ is the hypothesized difference
    - $s_i^2$ is the sample variance for group $i$
    - $n_i$ is the sample size of group $i$    
    - $\text{df} = \text{min}(n_1-1, n_2-1)$  

## Statistical Inference on Two Means (`R`)

- We will use the `independent_means` function from `library(ssstats)` to perform statistical inference.

- Generic syntax:

```{r}
#| eval: false
dataset_name |> independent_means( # <1>
  outcome = outcome_variable, # <2>
  grouping = grouping_variable, # <3>
  alternative = "two" | "less" | "greater", # <4>
  mu = hypothesized_difference, # <5>
  alpha = alpha_level # <6>
  ) 
```
1. Pipes your dataset into `independent_means()`.
2. `outcome`: the continuous variable you want to compare between the two groups.
3. `grouping`: the categorical variable that splits your data into the two groups being compared.
4. `alternative`: the direction of the test; default is `two`.
5. `mu`: the hypothesized difference between the two groups under $H_0$; default is 0.
6. `alpha`: the significance level; default is 0.05.

## Example 1

- Okonkwo pings you on your first morning back from vacation: "Patrol runs early shifts while Desk runs later ones and I keep hearing officers gripe about who's more rested. Tell me if there's actually a difference in how much sleep our two groups are getting."

- We will use the `zootopia` dataset to answer the following question:

    - Is there a difference in average nightly sleep between Patrol and Desk officers?

- First, we need to describe the data.

    - We start with point and interval estimates.
    
- Then, we will perform a hypothesis test to answer the question.

## Example 1
    
- Point and interval estimates:

```{r}
zootopia |> independent_means(outcome = sleep_hours,
                              grouping = department)
```

## Example 1

- Then, we will perform a hypothesis test to answer the question.    

    - Is there a difference in average nightly sleep between Patrol and Desk officers?
    
- How do we translate this into hypotheses?  

    - $H_0: \mu_{\text{Patrol}} = \mu_{\text{Desk}}$
    - $H_1: \mu_{\text{Patrol}} \ne \mu_{\text{Desk}}$
    
- For the `independent_means()` function, we will set 

    - `mu = 0`
    - `alternative = "two"`.  

## Example 1

- Is there a difference in average nightly sleep between Patrol and Desk officers?

```{r}
zootopia |> independent_means(outcome = sleep_hours,
                              grouping = department,
                              mu = 0,
                              alternative = "two")
```
    
## Example 1

- **Hypotheses**

    - $H_0: \mu_{\text{Desk}} = \mu_{\text{Patrol}}$
    - $H_1: \mu_{\text{Desk}} \ne \mu_{\text{Patrol}}$
    
- **Test Statistic and *p*-Value**

    - $t_0 = -1.737$
    - $p = 0.097$
    
- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Fail to reject $H_0$. There is not sufficient evidence to suggest that average sleep differs between Patrol and Desk officers.
    
## Example 2

- Okonkwo's back with another one: "Word around the precinct is that Patrol officers have more access to the public, so it's naturally easier to run their side hustles. Let's see what the data says -- do Patrol officers earn more than Desk officers?"

- We will use the `zootopia` dataset to answer the following question:

    - Do patrol officers earn more than desk officers?

- First, we need to describe the data.

    - We start with point and interval estimates.
    
- Then, we will perform a hypothesis test to answer the question.

## Example 2
    
- Point and interval estimates:

```{r}
zootopia |> independent_means(outcome = hustle_earnings,
                              grouping = department)
```

## Example 2

- Then, we will perform a hypothesis test to answer the question.    

    - Do patrol officers earn more than desk officers?
    
- How do we translate this into hypotheses?  

    - $H_0: \mu_{\text{Desk}} \ge \mu_{\text{Patrol}}$
    - $H_1: \mu_{\text{Desk}} < \mu_{\text{Patrol}}$
    
- For the `independent_means()` function, we will set 

    - `mu = 0`
    - `alternative = "greater"`.  

## Example 2

- Do patrol officers earn more than desk officers?

```{r}
zootopia |> independent_means(outcome = hustle_earnings,
                              grouping = department,
                              mu = 0,
                              alternative = "less")
```
    
## Example 2

- **Hypotheses**

    - $H_0: \mu_{\text{Desk}} \ge \mu_{\text{Patrol}}$
    - $H_1: \mu_{\text{Desk}} < \mu_{\text{Patrol}}$
    
- **Test Statistic and *p*-Value**

    - $t_0 = -0.917$
    - $p = 0.186$
    
- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Fail to reject $H_0$. There is not sufficient evidence to suggest that patrol officers earn more in side hustles than desk officers.

## Assumption on Two-Sample *t*

- We assume that the data has an approximate mound-shaped and symmetric distribution.

$$y_{11}, y_{12}, \ldots, y_{1n_1} \sim N(\mu_1, \sigma_1^2)$$
$$y_{21}, y_{22}, \ldots, y_{2n_2} \sim N(\mu_2, \sigma_2^2)$$

- We will use a quantile-quantile plot (Q-Q plot) to check the assumption of normality.

    - We will also eyeball a histogram to further check the assumption of normality.

## Checking Normality (`R`)

- We will use the `independent_qq()` function from `library(ssstats)` to assess normality.

```{r}
#| eval: false
dataset_name |> independent_qq(continuous = continuous_variable,
                               grouping = grouping_variable)
```

- This will provide the the Q-Q plots and the histograms for the two independent groups under consideration.


## A Note About Homoskedasticity

- Recall the assumptions,

$$y_{11}, y_{12}, \ldots, y_{1n_1} \sim N(\mu_1, \sigma_1^2)$$
$$y_{21}, y_{22}, \ldots, y_{2n_2} \sim N(\mu_2, \sigma_2^2)$$

- Note that we are not assuming $\sigma_1^2 = \sigma_2^2$.

    - The `independent_means()` function assumes $\sigma_1^2 \ne \sigma_2^2$.

## Example 1

- Let's check the normality assumption for the sleep data.

```{r}
#| fig-align: 'center'
zootopia |> independent_qq(outcome = sleep_hours,
                           grouping = department)
```

## Example 2

- Let's check the normality assumption for the side hustle earnings data.

```{r}
#| fig-align: 'center'
zootopia |> independent_qq(outcome = hustle_earnings,
                           grouping = department)
```


## Confidence Interval for $M_1-M_2$

- The confidence interval is built on every possible pairwise difference between the two groups.

    - We examine the $n_1 \times n_2$ differences between the observations.
    
        - i.e., We take each observation in Group 1 and subtract each observation in Group 2.
      
- The median of that pool of differences is the Hodges–Lehmann estimate.

    - It estimates the typical *shift* between the two groups.
    
- The CI is then found by

    - ranking that entire pool of $n_1 \times n_2$ differences from smallest to largest,
    
    - then we pick two values from that ranked list as the lower and upper bounds.

## Hypothesis Test for $M_1-M_2$

**Hypotheses**

::: {.hyp-columns .columns}

::: {.column width="30%"}
**Two-Tailed**

$H_0: M_1 - M_2 = M_0$

$H_1: M_1 - M_2 \ne M_0$
:::

::: {.column width="30%"}
**Left-Tailed**

$H_0: M_1 - M_2 \ge M_0$  

$H_1: M_1 - M_2 < M_0$
:::

::: {.column width="30%"}
**Right-Tailed**

$H_0: M_1 - M_2 \le M_0$

$H_1: M_1 - M_2 > M_0$
:::

:::

- $M_i$ is the population median for group $i$, and

- $M_0$ is the hypothesized difference between the groups

## Hypothesis Test for $M_1-M_2$

**Test Statistic**

$$
T_0 = \sum R_{\text{1}} - \frac{n_1(n_1+1)}{2}
$$

- where
    - $\sum R_1$ is the sum of the ranks for the  first group
    - $n_1$ is the sample size of the first group 

## Statistical Inference on Two Medians (`R`)

- We will use the `independent_medians` function from `library(ssstats)` to perform statistical inference.

- Generic syntax:

```{r}
#| eval: false
dataset_name |> independent_medians( # <1>
  outcome = outcome_variable, # <2>
  grouping = grouping_variable, # <3>
  alternative = "two" | "less" | "greater", # <4>
  m = hypothesized_difference, # <5>
  alpha = alpha_level # <6>
  ) 
```
1. Pipes your dataset into `independent_medians()`.
2. `outcome`: the continuous variable you want to compare between the two groups.
3. `grouping`: the categorical variable that splits your data into the two groups being compared.
4. `alternative`: the direction of the test; default is `two`.
5. `m`: the hypothesized difference between the two groups under $H_0$; default is 0.
6. `alpha`: the significance level; default is 0.05.

## Example 1

- Then, we will perform a hypothesis test to answer the question.    

    - Is there a difference in average (median) nightly sleep between Patrol and Desk officers?
    
- How do we translate this into hypotheses?  

    - $H_0: M_{\text{Patrol}} = M_{\text{Desk}}$
    - $H_1: M_{\text{Patrol}} \ne M_{\text{Desk}}$
    
- For the `independent_medians()` function, we will set 

    - `m = 0`
    - `alternative = "two"`.  

## Example 1

```{r}
zootopia |> independent_medians(outcome = sleep_hours,
                                grouping = department)
```

## Example 1

- **Hypotheses**

    - $H_0: M_{\text{Desk}} = M_{\text{Patrol}}$
    - $H_1: M_{\text{Desk}} \ne M_{\text{Patrol}}$
    
- **Test Statistic and *p*-Value**

    - $T_0 = 68$
    - $p = 0.092$
    
- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Fail to reject $H_0$. There is not sufficient evidence to suggest that median sleep differs between Patrol and Desk officers.

## Example 2

- Then, we will perform a hypothesis test to answer the question.    

    - Do patrol officers earn more than desk officers?
    
- How do we translate this into hypotheses?  

    - $H_0: M_{\text{Desk}} \ge M_{\text{Patrol}}$
    - $H_1: M_{\text{Desk}} < M_{\text{Patrol}}$
    
- For the `independent_means()` function, we will set 

    - `mu = 0`
    - `alternative = "greater"`.  

## Example 2

```{r}
zootopia |> independent_medians(outcome = hustle_earnings,
                              grouping = department, alternative = "less")
```

## Example 2

- **Hypotheses**

    - $H_0: M_{\text{Desk}} \ge M_{\text{Patrol}}$
    - $H_1: M_{\text{Desk}} < M_{\text{Patrol}}$
    
- **Test Statistic and *p*-Value**

    - $T_0 = 63$
    - $p = 0.029$
    
- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Reject $H_0$. There is sufficient evidence to suggest that median side hustle earnings are greater among those on patrol.


## Wrap Up

- Using a $t$-test to examine the difference between two independent means ($\mu_1-\mu_2$) requires normality.
    
    - We check normality with a quantile-quantile plot (`independent_qq()`).

- The analysis you choose should match the needed method:

    - If we meet normality (`independent_means()`):
    
        - Interval estimate: CI for $\mu_1-\mu_2$
        - One-sample $t$-test for $\mu_1-\mu_2$
        
    - If we do not meet normality (`independent_medians()`):
    
        - Interval estimate: CI for $M_1-M_2$
        - One-sample sign test for $M_1-M_2$

