---
title: "Dependent<br>Means and Medians"
execute:
  echo: true
  warning: false
  message: false
  error: true
format:
  revealjs:
    theme: simple
    css: style_lecture.css
    include-after-body: headers.html
    self-contained: true
    slide-number: true
    center: false
    width: 1600
    height: 900
    df-print: kable
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
    code-annotations: hover
editor: source
---

## Introduction
    
- In the last module, we reviewed statistical inference for one sample.
    
- In this module, we will focus on tests for two samples.
    - Two-sample independent means ($\mu_1-\mu_2$)
    - Two-sample dependent means ($\mu_d$)
    - Two-sample proportions ($\pi_1-\pi_2$)
    - Two-sample variances ($\sigma^2_1/\sigma^2_2$)
    
- In this lecture: two-sample dependent means ($\mu_d$) 

## Introduction

- So far in this module, we have focused on independent samples.
    - Two-sample means, independent data ($\mu_1-\mu_2$).
    - Two-sample medians, independent data ($M_1-M_2$).

- In this lecture, we shift to dependent (paired) samples.
    - Two-sample means, dependent data ($\mu_d$)
    - Two-sample medians, dependent data ($M_d$)

- The data is *paired* because we can link the observations.

    - The two measurements come from the *same* subject.
    
## New Dataset!

- You're still the ZPD's data analyst, but Deputy Chief Okonkwo has a different kind of question for you this time.

    - Before now, every question asked involved comparing different groups of recruits

    - Now, the questions are about the same recruits, measured twice.

- Okonkwo wants to know whether things are *changing*:

    - Did coaching help? 
    - Is fitness improving? 
    - Did new software actually speed things up? 

## `R` Setup 

```{r}
library(tidyverse)
library(ssstats)
```

- Our new Zootopia data,

```{r}
zootopia_paired <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/2-zootopia-paired.csv")
```

```{r}
#| echo: false

head(zootopia_paired, n = 3)
```

## Why Not Independent?

- If we run `independent_means()` on either the "before" column or the "after" column, we are ignoring the pairing.

- Because the same recruit is measured twice, their two scores are correlated.

    - A recruit who scored high the first time will likely also score high the second time.

- Pairing lets us focus on *within-subject change*.

## Dependent Data

- For each subject $i$, we calculate one difference score:

$$d_i = y_{i1} - y_{i2}$$

- All of our inference is now built on this *single column of differences*, not the original columns.

    - i.e., This means dependent-samples inference is really just one-sample inference (mean or median), applied to $d$.

## Confidence Interval for $\mu_d$

- $(1-\alpha)100\%$ CI for a population mean difference, $\mu_d$

$$\bar{d} \pm t_{\alpha/2, n-1} \frac{s_d}{\sqrt{n}}$$

- $\bar{d}$ is the mean of the differences, $s_d$ is the standard deviation of the differences, and $n$ is the number of **pairs** (not the number of raw observations).

- $t$ has $n-1$ degrees of freedom (df).

## Hypothesis Test for $\mu_d$

**Hypotheses**

::: {.hyp-columns .columns}

::: {.column width="30%"}
**Two-Tailed**

$H_0: \mu_d = \mu_0$

$H_1: \mu_d \ne \mu_0$
:::

::: {.column width="30%"}
**Left-Tailed**

$H_0: \mu_d \ge \mu_0$

$H_1: \mu_d < \mu_0$
:::

::: {.column width="30%"}
**Right-Tailed**

$H_0: \mu_d \le \mu_0$

$H_1: \mu_d > \mu_0$
:::

:::

- $\mu_d$ is the population mean of the differences,

    - Note that you must specify what $\mu_d$ is. e.g., $\mu_1-\mu_2$

- $\mu_0$ is the hypothesized value of that mean difference (often 0, for "no change").

## Hypothesis Test for $\mu_d$

**Test Statistic:**

$$
t_0 = \frac{\bar{d}-\mu_0}{{\dfrac{s_d}{\sqrt{n}}}} \sim t_{n-1}
$$

- where
    - $\bar{d}$ is the mean of the differences
    - $\mu_0$ is the hypothesized mean difference
    - $s_d$ is the sample standard deviation of the differences
    - $n$ is the number of pairs
    - df $= n - 1$

## Statistical Inference on Dependent Means (`R`)

- We will use the `dependent_means()` function from `library(ssstats)` to perform statistical inference.

- Generic syntax:

```{r}
#| eval: false
dataset_name |> dependent_means( # <1>
  col1 = first_measurement, # <2>
  col2 = second_measurement, # <3>
  alternative = "two" | "less" | "greater", # <4>
  mu = hypothesized_difference, # <5>
  alpha = alpha_level # <6>
  )
```
1. Pipes your dataset into `dependent_means()`.
2. `col1`: the first paired measurement.
3. `col2`: the second paired measurement. 
4. `alternative`: the direction of the test; default is `two`.
5. `mu`: the hypothesized mean difference under $H_0$; default is 0.
6. `alpha`: the significance level; default is 0.05.

- Remember that `R` will subtract `col1 - col2`.

## Example 1

- Okonkwo calls you in: "We started pulling recruits with low written-exam scores into a short remedial coaching session before their retake. I want proof this is worth the department's time -- did coaching actually raise their scores?"

- We will use the `zootopia_paired` dataset to answer the following question:

    - Did recruits' exam scores improve after the coaching session?

- We'll work with `exam_score_pre` and `exam_score_post` for the same recruits.

## Example 1

- Point and interval estimates:

```{r}
zootopia_paired |> dependent_means(col1 = exam_score_post,
                                   col2 = exam_score_pre)
```

## Example 1

- Did recruits' exam scores improve after coaching?

- How do we translate this into hypotheses?

    - We defined the difference as `exam_score_post - exam_score_pre`.
    - If coaching helped, scores went up after coaching, so the difference should be *positive*.
    
- **Hypotheses:**    

    - $H_0: \mu_d \le 0$, where $\mu_d = \mu_{\text{post}} - \mu_{\text{pre}}$
    - $H_1: \mu_d > 0$

- For the `dependent_means()` function, we will set

    - `mu = 0`
    - `alternative = "greater"`.

## Example 1

- Did recruits' exam scores improve after coaching?

```{r}
zootopia_paired |> dependent_means(col1 = exam_score_post,
                                   col2 = exam_score_pre,
                                   mu = 0,
                                   alternative = "greater")
```

## Example 1

- **Hypotheses**

    - $H_0: \mu_d \le 0$, where $\mu_d = \mu_{\text{post}} - \mu_{\text{pre}}$
    - $H_1: \mu_d > 0$

- **Test Statistic and *p*-Value**

    - $t_0 = 13.461$ 
    - $p < 0.001$ 

- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Reject $H_0$. There is sufficient evidence to suggest that the exam scores improved after coaching.

## Example 2

- Okonkwo again: "Recruits go through a fitness assessment when they start, and we test them again at the 6-month mark. Command wants to know -- overall, is fitness actually changing over those six months, or are we just imagining a trend?"

- We will use the `zootopia_paired` dataset to answer the following question:

    - Does recruit fitness change between intake and the 6-month mark?

- We'll work with `fitness_score_intake` and `fitness_score_6mo`.

## Example 2

- Point and interval estimates:

```{r}
zootopia_paired |> dependent_means(col1 = fitness_score_6mo,
                                   col2 = fitness_score_intake)
```

## Example 2

- Does recruit fitness change between intake and the 6-month mark?

- How do we translate this into hypotheses?

    - No direction or hypothesized value indicated.
    
- **Hypotheses:**    

    - $H_0: \mu_d = 0$, where $\mu_d = \mu_{\text{6 mo}} - \mu_{\text{intake}}$
    - $H_1: \mu_d \ne 0$

- For the `dependent_means()` function, we will set

    - `mu = 0`
    - `alternative = "two"`.

## Example 2

- Does recruit fitness change between intake and the 6-month mark?

```{r}
zootopia_paired |> dependent_means(col1 = fitness_score_6mo,
                                   col2 = fitness_score_intake,
                                   mu = 0,
                                   alternative = "two")
```

## Example 2

- **Hypotheses**

    - $H_0: \mu_d = 0$, where $\mu_d = \mu_{\text{6 mo}} - \mu_{\text{intake}}$
    - $H_1: \mu_d \ne 0$

- **Test Statistic and *p*-Value**

    - $t_0 = 0.70$
    - $p = 0.490$

- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Fail to reject $H_0$. There is not sufficient evidence to suggest that there is a change in fitness score within the first 6 months.

## Assumption on Paired *t*

- Just like the one-sample case, we assume the *differences* have an approximate mound-shaped, symmetric distribution.

$$d_1, d_2, \ldots, d_n \sim N(\mu_d, \sigma_d^2)$$

- Notice this assumption is about $d$, the differences, not about `col1` and `col2` individually.

    - The original two columns can each look however they look; it's the differences that need to be approximately normal.

- We will use a quantile-quantile (Q-Q) plot to check this, plus a histogram of the differences.

## Checking Normality (`R`)

- We will use the `dependent_qq()` function from `library(ssstats)` to assess normality.

```{r}
#| eval: false
dataset_name |> dependent_qq(col1 = first_measurement,
                             col2 = second_measurement)
```

- This produces one Q-Q plot and one histogram for the difference ($d_i$).

## Example 1

- Let's check the normality assumption for the exam score differences.

```{r}
#| fig-align: 'center'
zootopia_paired |> dependent_qq(col1 = exam_score_pre,
                                col2 = exam_score_post)
```

## Example 2

- Let's check the normality assumption for the fitness score differences.

```{r}
#| fig-align: 'center'
zootopia_paired |> dependent_qq(col1 = fitness_score_intake,
                                col2 = fitness_score_6mo)
```

## Confidence Interval for $M_d$

- Just like the independent-samples case, the CI for a median difference is built from ranks.

- Instead of pairwise differences *between* two groups, we now work with *averages of pairs of differences* within the same set of $d_i$'s.

    - For every pair of differences $d_i$ and $d_j$, compute the Welsh average, 

$$\dfrac{d_i + d_j}{2}$$

- The *median* of these averages is the Hodges-Lehmann estimate of the median difference.

- The CI is then found by ranking the Walsh averages and selecting upper and lower bounds.

## Ranking Paired Data

- The *differences* are ranked based on the <u>absolute value</u> of the difference.

- Then, ranks can be identified as "positive" or "negative" based on the original direction of the difference.

| *X* | *Y* | *D* | \|*D*\| | Rank |
|-----|-----|-----|---------|------|
| 5   | 8   | -3  | 3       | $-$ 1.5 |
| 8   | 5   | 3   | 3       | $+$ 1.5  |
| 4   | 4   | 0   | 0       | ---------   |

## Hypothesis Test for $M_d$

**Hypotheses**

::: {.hyp-columns .columns}

::: {.column width="30%"}
**Two-Tailed**

$H_0: M_d = M_0$

$H_1: M_d \ne M_0$
:::

::: {.column width="30%"}
**Left-Tailed**

$H_0: M_d \ge M_0$

$H_1: M_d < M_0$
:::

::: {.column width="30%"}
**Right-Tailed**

$H_0: M_d \le M_0$

$H_1: M_d > M_0$
:::

:::

- $M_d$ is the population median of the differences, 

    - Note that you must specify what $M_d$ is. e.g., $M_1-M_2$

- $M_0$ is the hypothesized median difference (often 0).

## Hypothesis Test for $M_d$

**Test Statistic**

$$
T_0 =
\begin{cases}
  R_+ = \text{sum of positive ranks} & \text{if left-tailed} \\
  R_- = \text{sum of negative ranks} & \text{if right-tailed} \\
  \min(R_+, R_-) & \text{if two-tailed}
\end{cases}
$$

## Statistical Inference on Dependent Medians (`R`)

- We will use the `dependent_medians()` function from `library(ssstats)` to perform statistical inference.

- Generic syntax:

```{r}
#| eval: false
dataset_name |> dependent_medians( # <1>
  col1 = first_measurement, # <2>
  col2 = second_measurement, # <3>
  alternative = "two" | "less" | "greater", # <4>
  m = hypothesized_difference, # <5>
  alpha = alpha_level # <6>
  )
```
1. Pipes your dataset into `dependent_medians()`.
2. `col1`: the first paired measurement.
3. `col2`: the second paired measurement. **The function calculates `col1 - col2`, so order matters.**
4. `alternative`: the direction of the test; default is `two`.
5. `m`: the hypothesized median difference under $H_0$; default is 0.
6. `alpha`: the significance level; default is 0.05.

## Example 1

- Recall the research question: 

    - Did recruits' exam scores improve after coaching?

- **Hypotheses:**    

    - $H_0: M_d \le 0$, where $M_d = M_{\text{post}} - M_{\text{pre}}$
    - $H_1: M_d > 0$

- For the `dependent_medians()` function, we will set

    - `mu = 0`
    - `alternative = "greater"`.

## Example 1

```{r}
zootopia_paired |> dependent_medians(col1 = exam_score_post,
                                     col2 = exam_score_pre,
                                     alternative = "greater")
```

## Example 1

- **Hypotheses**

    - $H_0: M_d \le 0$, where $M_d = M_{\text{post}} - M_{\text{pre}}$
    - $H_1: M_d > 0$

- **Test Statistic and *p*-Value**

    - $T_0 = 465$ 
    - $p < 0.001$ 

- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Reject $H_0$. There is sufficient evidence to suggest that the exam scores improved after coaching.

## Example 2

- Recall: the fitness score differences looked questionable on the Q-Q plot -- some recruits improved, some plateaued, some backslid.

    - This is exactly the situation where we don't want to lean on a mean-based test.

- How do we translate the question into hypotheses this time?

    - $H_0: M_d = 0$
    - $H_1: M_d \ne 0$

- For the `dependent_medians()` function, we will set

    - `m = 0`
    - `alternative = "two"`.

## Example 2

- Recall the research question:

    - Does recruit fitness change between intake and the 6-month mark?

```{r}
zootopia_paired |> dependent_medians(col1 = fitness_score_6mo,
                                     col2 = fitness_score_intake)
```

## Example 2 (Revisited)

- **Hypotheses**

    - $H_0: M_d = 0$
    - $H_1: M_d \ne 0$

- **Test Statistic and *p*-Value**

    - $T_0 = 272$
    - $p = 0.406$ 

- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Fail to reject $H_0$. There is not sufficient evidence to suggest that fitness changes between intake and their 6 month check in.
    
## Wrap Up

- Dependent (paired) data comes from the same subject measured twice.

    - We always work with the differences, $d_i = \text{col1}_i - \text{col2}_i$.

- The analysis you choose should match the shape of the differences:

    - If the differences meet normality (`dependent_means()`):

        - Interval estimate: CI for $\mu_d$
        - Paired $t$-test for $\mu_d$

    - If the differences do not meet normality (`dependent_medians()`):

        - Interval estimate: CI for $M_d$ (built from Walsh averages)
        - Wilcoxon signed-rank test for $M_d$

