---
title: "Two-Sample Proportions"
execute:
  echo: true
  warning: false
  message: false
  error: true
format:
  revealjs:
    theme: simple
    css: style_lecture.css
    include-after-body: headers.html
    self-contained: true
    slide-number: true
    center: false
    width: 1600
    height: 900
    df-print: kable
    html-math-method: katex
    code-fold: false
    code-tools: false
    code-annotations: hover
    incremental: false
editor: source
---

## Introduction
    
- In the last module, we reviewed statistical inference for one sample.
    
- In this module, we will focus on tests for two samples.
    - Two-sample independent means ($\mu_1-\mu_2$)
    - Two-sample dependent means ($\mu_d$)
    - Two-sample proportions ($\pi_1-\pi_2$)
    - Two-sample variances ($\sigma^2_1/\sigma^2_2$)
    
- In this lecture: two-sample proportions ($\pi_1-\pi_2$) 

## `R` Setup 

```{r}
library(tidyverse)
library(ssstats)
```

- Both examples today use a new dataset for two ZPD precincts, `Meadowbrook` and `Coldharbor`, each with 60 officers who just went through recertification.

```{r}
zootopia <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/3-zootopia-props.csv")
```

```{r}
#| echo: false

head(zootopia, n = 3)
```

## Confidence Interval for $\pi_1-\pi_2$

- (1--$\alpha$)100% CI for a difference between population proportions, $\pi_1-\pi_2$

$$
(\hat{p}_1 - \hat{p}_2) \pm z_{\alpha/2} \sqrt{\dfrac{\hat{p}_1(1-\hat{p}_1)}{n_1} + \dfrac{\hat{p}_2(1-\hat{p}_2)}{n_2}}
$$

## Hypothesis Test for $\pi_1-\pi_2$

- **Hypotheses:**

::: {.hyp-columns .columns}

::: {.column width="30%"}
**Two-Tailed**

$H_0: \ \pi_1-\pi_2=\pi_0$

$H_1: \ \pi_1-\pi_2 \ne \pi_0$
:::

::: {.column width="30%"}
**Left-Tailed**

$H_0: \ \pi_1-\pi_2 \ge \pi_0$

$H_1: \ \pi_1-\pi_2 < \pi_0$
:::

::: {.column width="30%"}
**Right-Tailed**

$H_0: \ \pi_1-\pi_2 \le \pi_0$

$H_1: \ \pi_1-\pi_2 > \pi_0$
:::

:::

- $\pi_i$ is the population proportion for group $i$

- $\pi_0$ is the hypothesized value of the difference in proportions (often 0, for "no change").
    
## Hypothesis Test for $\pi_1-\pi_2$

- **Test Statistic**

$$z_0 = \frac{\left( \hat{p}_1 - \hat{p}_2 \right)- d_0}{\sqrt{\hat{p}\left(1-\hat{p}\right)\left( \frac{1}{n_1}+\frac{1}{n_2} \right)}}$$

- where 
    - $\hat{p}_i$ is the sample proportion for group $i$,
    - $n_i$ is the sample size for group $i$, and
    
## Hypothesis Test for $\pi_1-\pi_2$

- **Test Statistic**

$$z_0 = \frac{\left( \hat{p}_1 - \hat{p}_2 \right)- d_0}{\sqrt{\hat{p}\left(1-\hat{p}\right)\left( \frac{1}{n_1}+\frac{1}{n_2} \right)}}$$

- where     
    - $\hat{p}$ is the pooled proportion, given by
    
$$\hat{p} = \frac{x_1+x_2}{n_1+n_2}$$

## Statistical Inference on Two Proportions (`R`)

- We will use the `independent_proportions()` function from `library(ssstats)` to perform statistical inference.

- Generic syntax:

```{r}
#| eval: false
dataset_name |> independent_proportions(outcome = binary_variable, # <1>
                                        grouping = grouping_variable, # <2>
                                        event = "Name of Event", # <3>
                                        p = hypothesized_value, # <4>
                                        alternative = "two" | "less" | "greater", # <5>
                                        alpha = specified_alpha)  # <6>
```
1. The binary outcome column — the variable that gets sorted into "success"/"not success."
2. The grouping column that splits your data into the two independent samples being compared. Must have exactly two levels.
3. The value of `binary_variable` that counts as a "success" — this defines what your proportions are proportions *of*.
4. The hypothesized difference between the two groups' true proportions under $H_0$; default is `0`.
5. The direction of $H_1$; default is `"two"`
6. The significance level $\alpha$, which also sets the CI's confidence level at $(1-\alpha)100\%$; default is `0.05`.

## Example 1

- Meadowbrook and Coldharbor just finished the same taser recertification course, same instructors, same exam. Deputy Chief Okonkwo wants to verify that Meadowbrook and Coldharbor recertification rates look the same.

- We will use the `zootopia` dataset to answer the following question:

    - Does the taser recertification pass rate differ between Meadowbrook and Coldharbor?

- We'll work with `taser_recert` and `precinct`.

## Example 1

- Point and interval estimates:

```{r}
zootopia |> independent_proportions(outcome = taser_recert,
                                    grouping = precinct,
                                    event = "Pass")
```

## Example 1

- Does the taser recertification pass rate differ between Meadowbrook and Coldharbor?

- How do we translate this into hypotheses?

    - No direction or hypothesized value given.
    
- **Hypotheses:**    

    - $H_0: \pi_{\text{MB}} = \pi_{\text{CH}}$
    - $H_1: \pi_{\text{MB}} \ne \pi_{\text{CH}}$

- For the `independent_proportion()` function, we will set

    - `p = 0`
    - `alternative = "two"`.

## Example 1

- Does the taser recertification pass rate differ between Meadowbrook and Coldharbor?

```{r}
zootopia |> independent_proportions(outcome = taser_recert,
                                    grouping = precinct,
                                    event = "Pass")
```

## Example 1

- **Hypotheses**

    - $H_0: \pi_{\text{MB}} = \pi_{\text{CH}}$
    - $H_1: \pi_{\text{MB}} \ne \pi_{\text{CH}}$

- **Test Statistic and *p*-Value**

    - $t_0 = -0.58$ 
    - $p = 0.559$ 

- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Fail to reject $H_0$. There is not sufficient evidence to suggest that the recertification rate is different between the two precincts.

## Example 2

- Coldharbor got a new firearms range last year... the one Meadowbrook's been asking for forever. If it's actually helping, Coldharbor's requalification numbers should be better. Let's see if the budget was worth it.

- We will use the `zootopia` dataset to answer the following question:

    - Does Coldharbor have a higher firearms requalification pass rate than Meadowbrook?
    
- We'll work with `firearms_requal` and `precinct`.    


## Example 2

- Point and interval estimates:

```{r}
zootopia |> independent_proportions(outcome = firearms_requal,
                                    grouping = precinct,
                                    event = "Pass")
```

## Example 2

- Does Coldharbor have a higher firearms requalification pass rate than Meadowbrook?

- How do we translate this into hypotheses?

    - Direction given: "higher"
    
- **Hypotheses:**    

    - $H_0: \pi_{\text{CH}} \le \pi_{\text{MB}}$
    - $H_1: \pi_{\text{CH}} > \pi_{\text{MB}}$

- For the `independent_proportion()` function, we will set

    - `p = 0`
    - `alternative = "greater"`.

## Example 2

- Does Coldharbor have a higher firearms requalification pass rate than Meadowbrook?

```{r}
zootopia |> independent_proportions(outcome = firearms_requal,
                                    grouping = precinct,
                                    event = "Pass",
                                    alternative = "greater")
```

## Example 2

- **Hypotheses**

    - $H_0: \pi_{\text{CH}} \le \pi_{\text{MB}}$
    - $H_1: \pi_{\text{CH}} > \pi_{\text{MB}}$

- **Test Statistic and *p*-Value**

    - $z_0 = 2.11$ 
    - $p = 0.018$ 

- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Reject $H_0$. There is sufficient evidence to suggest that the requalification rate is higher in Coldharbor as compared to Meadowbrook.


## Wrap Up

- Independent (two-sample) proportion data comes from two separate, unrelated groups.

    - We work with the difference in sample proportions, $\hat{p}_1 - \hat{p}_2$.
    
- Next: comparing two variances.