---
title: "One-Way ANOVA"
execute:
  echo: true
  warning: false
  message: false
  error: true
format:
  revealjs:
    theme: simple
    css: style_lecture.css
    include-after-body: headers.html
    self-contained: true
    slide-number: true
    center: false
    width: 1600
    height: 900
    df-print: kable
    html-math-method: katex
    code-fold: false
    code-annotations: hover
    code-tools: false
    incremental: false
editor: source
---

## Introduction

- Previous: one sample, two samples

- Now: more than two groups, continuous outcome

    - One-way ANOVA

    - Kruskal-Wallis

    - posthoc testing
    
- This lecture: one-way ANOVA    

## `R` Setup 

```{r}
library(tidyverse)
library(ssstats)
commute_times <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/4-2-commute_times.csv")
citations <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/4-4-citations.csv")
academy_scores <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/4-1-academy_scores.csv")
```
    
## Introduction: Analysis of Variance

- We have previously discussed testing the difference between two groups.

    - What about when there are three or more groups? 

- We will use a method called **an**alysis **o**f **va**riance (ANOVA).

    - This method partitions the variance of the outcome into variance due to the groups and variance due to "other" factors.

- Fun fact: the two-sample *t*-test is a special case of ANOVA.

    - If you perform ANOVA when comparing two means, you will obtain the same results as the two-sample *t*-test.    

## Analysis of Variance

- Consider comparing $k$ groups. 

<style>
  .anova-table {
    width: 100%;
    max-width: 1100px;
    margin: 30px auto;
    border-collapse: separate;
    border-spacing: 0;
    font-family: "Aptos", "Avenir Next", Arial, sans-serif;
    font-size: 26px;
    color: #263238;
    background: white;
    border-radius: 12px;
    overflow: hidden;
    box-shadow: 0 4px 18px rgba(0, 0, 0, 0.10);
  }

  .anova-table th {
    background: #263f5c;
    color: white;
    font-weight: 600;
    padding: 18px 20px;
    text-align: center;
    border-bottom: 4px solid #8fb3cc;
  }

  .anova-table th:first-child {
    width: 18%;
  }

  .anova-table th:nth-child(2) {
    width: 48%;
  }

  .anova-table th:nth-child(3),
  .anova-table th:nth-child(4) {
    width: 17%;
  }

  .anova-table td {
    padding: 14px 20px;
    text-align: center;
    border-bottom: 1px solid #d9e1e8;
  }

  .anova-table tbody tr:nth-child(even) {
    background: #f4f7fa;
  }

  .anova-table tbody tr:last-child td {
    border-bottom: none;
  }

  .anova-table .treatment {
    font-weight: 600;
    color: #263f5c;
  }

  .anova-table .observations {
    font-family: "Cambria Math", "STIX Two Math", serif;
    font-style: italic;
    letter-spacing: 1px;
  }

  .anova-table .summary {
    background: #edf3f7;
    font-family: "Cambria Math", "STIX Two Math", serif;
    font-style: italic;
    font-weight: 600;
  }

  .anova-table .dots {
    color: #78909c;
    font-style: normal;
  }

  .anova-table .symbol {
    font-family: "Cambria Math", "STIX Two Math", serif;
    font-style: italic;
  }

  .anova-table sub {
    font-size: 0.65em;
  }
</style>

<table class="anova-table">
  <thead>
    <tr>
      <th>Treatment<br>(Level)</th>
      <th>Observations</th>
      <th>Total</th>
      <th>Average</th>
    </tr>
  </thead>

  <tbody>
    <tr>
      <td class="treatment">1</td>
      <td class="observations">
        y<sub>11</sub>&nbsp;&nbsp;
        y<sub>12</sub>&nbsp;&nbsp;
        <span class="dots">⋯</span>&nbsp;&nbsp;
        y<sub>1n</sub>
      </td>
      <td class="summary">
        y<sub>1·</sub>
      </td>
      <td class="summary">
        ȳ<sub>1·</sub>
      </td>
    </tr>

    <tr>
      <td class="treatment">2</td>
      <td class="observations">
        y<sub>21</sub>&nbsp;&nbsp;
        y<sub>22</sub>&nbsp;&nbsp;
        <span class="dots">⋯</span>&nbsp;&nbsp;
        y<sub>2n</sub>
      </td>
      <td class="summary">
        y<sub>2·</sub>
      </td>
      <td class="summary">
        ȳ<sub>2·</sub>
      </td>
    </tr>

    <tr>
      <td class="treatment dots">⋮</td>
      <td class="observations dots">⋮</td>
      <td class="summary dots">⋮</td>
      <td class="summary dots">⋮</td>
    </tr>

    <tr>
      <td class="treatment">k</td>
      <td class="observations">
        y<sub>k1</sub>&nbsp;&nbsp;
        y<sub>k2</sub>&nbsp;&nbsp;
        <span class="dots">⋯</span>&nbsp;&nbsp;
        y<sub>kn</sub>
      </td>
      <td class="summary">
        y<sub>k·</sub>
      </td>
      <td class="summary">
        ȳ<sub>k·</sub>
      </td>
    </tr>
  </tbody>
</table>

## ANOVA Tables

- We will see our ANOVA results displayed in a table.

- Here is what we are working towards:

<table class="anova-table">
  <thead>
    <tr>
      <th>Source</th>
      <th>df</th>
      <th>SS</th>
      <th>MS</th>
      <th>F</th>
      <th>p-value</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td class="treatment">Treatment</td>
      <td class="summary">df<sub>Trt</sub></td>
      <td class="summary">SS<sub>Trt</sub></td>
      <td class="summary">MS<sub>Trt</sub></td>
      <td class="summary">F<sub>Trt</sub></td>
      <td class="summary"><em>p</em><sub>Trt</sub></td>
    </tr>
    <tr>
      <td class="treatment">Residuals</td>
      <td class="summary">df<sub>E</sub></td>
      <td class="summary">SS<sub>E</sub></td>
      <td class="summary">MS<sub>E</sub></td>
      <td></td>
      <td></td>
    </tr>
    <tr>
      <td class="treatment">Total</td>
      <td class="summary">df<sub>Tot</sub></td>
      <td class="summary">SS<sub>Tot</sub></td>
      <td></td>
      <td></td>
      <td></td>
    </tr>
  </tbody>
</table>

## ANOVA: Sums of Squares

- We are *partitioning the total variability* into different components. 

- The total sum of squares measures the overall variability in the data:

$$ \text{SS}_{\text{Tot}} = \sum_{i=1}^k \sum_{j=1}^n (y_{ij} - \bar{y}_{..})^2 $$

- and is partitioned into SS<sub>Trt</sub> and SS<sub>E</sub>:

$$\text{SS}_\text{Trt}=n \sum_{i=1}^k \left( \bar{y}_{i.}-\bar{y}_{..} \right)^2  \ \ \ \ \ \ \ \ \ \   \text{SS}_\text{E} = \sum_{i=1}^k \sum_{j=1}^n (y_{ij} - \bar{y}_{i.})^2$$

## ANOVA: Degrees of Freedom

- Next, we have the degrees of freedom. Each sum of square will have its own degrees of freedom.

- The total df:

$$\text{df}_{\text{Tot}} = n-1$$

- and is partitioned into df<sub>Trt</sub> and df<sub>E</sub>:

$$\text{df}_\text{Trt}=k - 1  \ \ \ \ \ \ \ \ \ \   \text{df}_\text{E} = n - k$$

## ANOVA: Mean Squares and Test Statistic

- To find the mean squares, we simply divide the sum of squares by the appropriate degrees of freedom,

$$\text{MS}_\text{Trt}=\frac{\text{SS}_{\text{Trt}}}{\text{df}_{\text{Trt}}} \ \ \ \ \ \ \ \ \ \   \text{MS}_\text{E}=\frac{\text{SS}_{\text{E}}}{\text{df}_{\text{E}}}$$

- Then, we can find our test statistic,

$$F_{\text{Trt}} = \frac{\text{MS}_{\text{Trt}}}{\text{MS}_{\text{E}}}$$

## ANOVA: Mean Squares and Test Statistic

- Note for the future: the MS and *F* can be generalized --

$$\text{MS}_\text{X}=\frac{\text{SS}_{\text{X}}}{\text{df}_{\text{X}}} \ \ \ \ \ \ \ \ \ \  F_{\text{X}} = \frac{\text{MS}_{\text{X}}}{\text{MS}_{\text{E}}}$$

- When constructing a test statistic in ANOVA, we always divide by MS<sub>E</sub>.

## ANOVA Tables

- Revisiting the ANOVA table,

<table class="anova-table">
  <thead>
    <tr>
      <th>Source</th>
      <th>df</th>
      <th>SS</th>
      <th>MS</th>
      <th>F</th>
      <th>p-value</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td class="treatment">Treatment</td>
      <td class="summary">df<sub>Trt</sub></td>
      <td class="summary">SS<sub>Trt</sub></td>
      <td class="summary">MS<sub>Trt</sub></td>
      <td class="summary">F<sub>Trt</sub></td>
      <td class="summary"><em>p</em><sub>Trt</sub></td>
    </tr>
    <tr>
      <td class="treatment">Residuals</td>
      <td class="summary">df<sub>E</sub></td>
      <td class="summary">SS<sub>E</sub></td>
      <td class="summary">MS<sub>E</sub></td>
      <td></td>
      <td></td>
    </tr>
    <tr>
      <td class="treatment">Total</td>
      <td class="summary">df<sub>Tot</sub></td>
      <td class="summary">SS<sub>Tot</sub></td>
      <td></td>
      <td></td>
      <td></td>
    </tr>
  </tbody>
</table>

- Note the relationships:

    - SS<sub>Trt</sub> + SS<sub>E</sub> = SS<sub>Tot</sub>
    - df<sub>Trt</sub> + df<sub>E</sub> = df<sub>Tot</sub>

## Hypothesis Test for $\mu_1 = \mu_2 = ... = \mu_k$

- **Hypotheses**

    - $H_0: \mu_1 = \mu_2 = ... = \mu_k$
    - $H_1:$ at least one mean is different.
    
- **Test Statistic:**

    - $F_0 = \text{MS}_{\text{Trt}} / \text{MS}_{\text{E}}$

## Statistical Inference on $k$ Means (`R`)

- We will use the `anovas()` function from `library(ssstats)` to construct the ANOVA table.

```{r}
#| eval: false
dataset_name |> anovas(outcome = continuous_variable, # <1>
                       grouping = grouping_variable, # <2>
                       alpha = alpha_level) # <3>
```
1. `outcome`: the continuous variable being compared across groups.
2. `grouping`: the categorical variable defining the groups (must have at least 2 levels).
3. `alpha`: the significance level; default is `0.05`.



## Example 1

- Okonkwo catches you between meetings: "Officers in Northgate keep grumbling that their commute is worse than everyone else's. Before I approve any relocation stipends, I want to know if commute time actually differs across our four districts."

```{r}
#| echo: false
commute_times |> head(n = 3)
```

- We will use the `commute_times` dataset to answer the following question:

    - Does commute time differ across the four districts?
    
## Example 1

- Does commute time differ across the four districts?

- How do we translate this into hypotheses?

    - No direction or hypothesized value.
    
    - We need to enumerate the number of means expressed in $H_0$.
    
- **Hypotheses:**    

    - $H_0: \mu_{\text{M}} = \mu_{\text{C}} = \mu_{\text{N}} = \mu_{\text{F}}$
    - $H_1:$ at least one district is different

## Example 1

- Does commute time differ across the four districts?

```{r}
commute_times |> anovas(outcome = commute_minutes,
                         grouping = district)
```

## Example 1

- **Hypotheses**

    - $H_0: \mu_{\text{M}} = \mu_{\text{C}} = \mu_{\text{N}} = \mu_{\text{F}}$
    - $H_1:$ at least one district is different

- **Test Statistic and *p*-Value**

    - $F_0 = 0.780$
    - $p = 0.507$

- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Fail to reject $H_0$. There is not sufficient evidence to suggest that there is a difference in commute times between the districts.    
    
## Example 2

- Okonkwo swings by your desk: "We just finished processing assessment scores for the incoming class. This cycle, recruits went through training at three different academies: Meadowbrook, Coldharbor, and Northgate. Command wants to know if it matters which academy a recruit trains at. Are scores different across the three?"

```{r}
#| echo: false
academy_scores |> head(n = 3)
```

- We will use the `academy_scores` dataset to answer the following question:

    - Do assessment scores differ across the three training academies?

## Example 2

- Do assessment scores differ across the three training academies (Meadowbrook, Coldharbor, and Northgate)?

- How do we translate this into hypotheses?

    - No direction or hypothesized value.
    
    - We need to enumerate the number of means expressed in $H_0$.
    
- **Hypotheses:**    

    - $H_0: \mu_{\text{M}} = \mu_{\text{C}} = \mu_{\text{N}}$
    - $H_1:$ at least one training academy is different

## Example 2

- Do assessment scores differ across the three training academies (Meadowbrook, Coldharbor, and Northgate)?

```{r}
academy_scores |> anovas(outcome = score,
                         grouping = academy)
```

## Example 2

- **Hypotheses**

    - $H_0: \mu_{\text{M}} = \mu_{\text{C}} = \mu_{\text{N}}$
    - $H_1:$ at least one training academy is different

- **Test Statistic and *p*-Value**

    - $F_0 = 5.898$
    - $p = 0.004$

- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Reject $H_0$. There is sufficient evidence to suggest that there is a difference in assessment scores between academies.    

## Wrap Up

- This lecture introduced one-way ANOVA.

    - One-way ANOVA examines the effect of a single factor.
    
    - (The next module introduces multiple factors!)
    
- One-way ANOVA only allows us to determine if there is an overall, or omnibus, effect.

    - It does *not* tell us about pairwise differences.
    
- In the future:

    - Assessing ANOVA assumptions
    - Nonparametric alternative
    - Posthoc testing