---
title: "ANOVA Assumptions"
execute:
  echo: true
  warning: false
  message: false
  error: true
format:
  revealjs:
    theme: simple
    css: style_lecture.css
    include-after-body: headers.html
    self-contained: true
    slide-number: true
    center: false
    width: 1600
    height: 900
    df-print: kable
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
    code-annotations: hover
editor: source
---

## Introduction

- Previous: one sample, two samples

- Now: more than two groups, continuous outcome

    - One-way ANOVA

    - Kruskal-Wallis

    - posthoc testing
    
- This lecture: ANOVA assumptions    

## `R` Setup 

```{r}
library(tidyverse)
library(ssstats)
commute_times <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/4-2-commute_times.csv")
citations <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/4-4-citations.csv")
academy_scores <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/4-1-academy_scores.csv")
```

## ANOVA Assumptions

- We can represent ANOVA with the following model:

$$y_{ij} = \mu + \tau_i + \varepsilon_{ij}$$

- where:

    - $y_{ij}$ is the $j^{\text{th}}$ observation in the $i^{\text{th}}$ group,
    - $\mu$ is the overall (grand) mean,
    - $\tau_i$ is the treatment effect for group $i$, and
    - $\varepsilon_{ij}$ is the error term for the $j^{\text{th}}$ observation in the $i^{\text{th}}$ group.
    
## ANOVA Assumptions

- We assume that the error term follows a normal distribution with mean 0 and a constant variance, $\sigma^2$. i.e.,

$$\varepsilon_{ij} \overset{\text{iid}}{\sim} N(0, \sigma^2)$$

- Very important note: **the assumption is on the error term** ($\varepsilon_{ij}$) and NOT on the outcome ($y_{ij}$)!

## ANOVA Assumptions

- We will use the residual (the difference between the observed value and the predicted value) to assess assumptions:

$$e_{ij} = y_{ij} - \hat{y}_{ij}$$

- Note that in the case of one-way ANOVA, the predicted value of $y$ is the *group mean*.

## ANOVA Assumptions: Graphical Assessment
    
- **Normality**: quantile-quantile plot

    - Should have points close to the 45$^\circ$ line
    - We will focus on the "center" portion of the plot
    
- **Variance**: scatterplot of the residuals against the predicted values

    - Should be "equal spread" between the groups
    - No "pattern"

## Graphical Assessment of Assumptions (`R`)

- We will use the `anova_check()` function from `library(ssstats)` to assess normality.

```{r}
#| eval: false
dataset_name |> anova_check(outcome = continuous_variable, # <1>
                       grouping = grouping_variable) # <2>
```
1. `outcome`: the continuous variable being compared across groups.
2. `grouping`: the categorical variable defining the groups (must have at least 2 levels).

- This will provide the the Q-Q plots and variance scatterplot necessary to assess the ANOVA assumptions.

## Example 1

- Recall our first question: does commute time differ across the four districts?

```{r}
commute_times |> anovas(outcome = commute_minutes,
                        grouping = district)
```

## Example 1

- Checking the assumptions,

```{r}
#| fig-align: 'center'
commute_times |> anova_check(outcome = commute_minutes,
                             grouping = district)
```

## Example 1

- If it's hard to "see" the qq plots, we can "zoom in" using `one_qq()`,

```{r}
#| fig-align: 'center'
commute_times |> filter(district == "Coldharbor") |> one_qq(commute_minutes)
```

## Example 1

- If it's hard to "see" the qq plots, we can "zoom in" using `one_qq()`,

```{r}
#| fig-align: 'center'
commute_times |> filter(district == "Fairview") |> one_qq(commute_minutes)
```

## Example 1

- If it's hard to "see" the qq plots, we can "zoom in" using `one_qq()`,

```{r}
#| fig-align: 'center'
commute_times |> filter(district == "Meadowbrook") |> one_qq(commute_minutes)
```

## Example 1

- If it's hard to "see" the qq plots, we can "zoom in" using `one_qq()`,

```{r}
#| fig-align: 'center'
commute_times |> filter(district == "Northgate") |> one_qq(commute_minutes)
```

## Example 1

- Final assessment:

    - Normality: fine
    
    - Variance: ehhhh...?

## Brown-Forsythe-Levine Test

- When we are not confident making a call graphically, we can formally check the variance.

    - The analytic approach:
    
        - Transform $y_{ij}$ by subtracting off its group median.
        - Run ANOVA on these transformed values.

- **Hypotheses:**

    - $H_0$: $\sigma_1^2=\sigma_2^2=\hdots=\sigma_k^2$
    - $H_1$: population variances are not all equal
    
- **Test Statistic:**

    - $F_0$ on transformed data

## Checking Variance (`R`)

- We will use the `multiple_variances()` function from `library(ssstats)` to formally test the equality of variances.

```{r}
#| eval: false
dataset_name |> multiple_variances(outcome = continuous_variable, # <1>
                                   grouping = grouping_variable, # <2>
                                   alpha = alpha_level) # <3>
```
1. `outcome`: the continuous variable being compared across groups.
2. `grouping`: the categorical variable defining the groups (must have at least 2 levels).
3. `alpha`: the significance level; default is `0.05`.

## Example 1

- Checking the variance,

```{r}
commute_times |> multiple_variances(outcome = commute_minutes,
                                    grouping = district)
```

## Example 1

- **Hypotheses**

    - $H_0: \sigma^2_{\text{M}} = \sigma^2_{\text{C}} = \sigma^2_{\text{N}} = \sigma^2_{\text{F}}$
    - $H_1:$ at least one district is different

- **Test Statistic and *p*-Value**

    - $F_0 = 0.129$
    - $p = 0.943$

- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Fail to reject $H_0$. There is not sufficient evidence to suggest that there is a difference in the variance of commute times between the districts.    

## Example 2

- Do assessment scores differ across the three training academies (Meadowbrook, Coldharbor, and Northgate)?

```{r}
academy_scores |> anovas(outcome = score,
                         grouping = academy)
```

## Example 2

- Looking at the assumptions graphically,

```{r}
#| fig-align: 'center'
academy_scores |> anova_check(outcome = score,
                              grouping = academy)
```

## Example 2

- Verifying the variance assumption,

```{r}
academy_scores |> multiple_variances(outcome = score,
                                     grouping = academy)
```

## Example 2

- **Hypotheses**

    - $H_0: \sigma^2_{\text{M}} = \sigma^2_{\text{C}} = \sigma^2_{\text{N}}$
    - $H_1:$ at least one district's academy is different

- **Test Statistic and *p*-Value**

    - $F_0 = 0.914$
    - $p = 0.404$

- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Fail to reject $H_0$. There is not sufficient evidence to suggest that there is a difference in the variance of assessment scores between the academies    

## Example 3

- Okonkwo's last one for the day: "Traffic enforcement wants budget for more radar units. Their pitch is that citation activity is way higher in some districts than others. Before I sign off, let's check the citation counts across the four districts."

- We will use the `citations` dataset to answer the following question:

    - Does the number of citations issued per officer differ across the four districts?

## Example 3

- Does the number of citations issued per officer differ across the four districts?

```{r}
#| fig-align: 'center'
citations |> anova_check(outcome = citations,
                         grouping = district)
```

## Example 3

- Does the number of citations issued per officer differ across the four districts?

```{r}
#| fig-align: 'center'
#| echo: false
citations |> filter(district == "Coldharbor") |> one_qq(citations)
```

## Example 3

- Does the number of citations issued per officer differ across the four districts?

```{r}
#| fig-align: 'center'
#| echo: false
citations |> filter(district == "Fairview") |> one_qq(citations)
```

## Example 3

- Verifying the variance assumption,

```{r}
citations |> multiple_variances(outcome = citations,
                                grouping = district)
```

## Example 3

- **Hypotheses**

    - $H_0: \sigma^2_{\text{M}} = \sigma^2_{\text{C}} = \sigma^2_{\text{N}} = \sigma^2_{\text{F}}$
    - $H_1:$ at least one district is different

- **Test Statistic and *p*-Value**

    - $F_0 = 4.237$
    - $p = 0.007$

- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Reject $H_0$. There is sufficient evidence to suggest that there is a difference in the variance of citations between the districts.   
    
## Wrap Up

- This lecture introduced assessing assumptions for one-way ANOVA.

    - Normality: graph only
    
    - Variance: graph and test available
    
- However, we have not yet discussed what to do about a broken assumption...    
    
- In the future:

    - Nonparametric alternative
    - Posthoc testing