---
title: "Kruskal-Wallis"
execute:
  echo: true
  warning: false
  message: false
  error: true
format:
  revealjs:
    theme: simple
    css: style_lecture.css
    include-after-body: headers.html
    self-contained: true
    slide-number: true
    center: false
    width: 1600
    height: 900
    df-print: kable
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
    code-annotations: hover
editor: source
---

## Introduction

- Previous: one sample, two samples

- Now: more than two groups, continuous outcome

    - One-way ANOVA

    - Kruskal-Wallis

    - posthoc testing
    
- This lecture: Kruskal-Wallis (nonparametric alternative)    

## `R` Setup 

```{r}
library(tidyverse)
library(ssstats)
commute_times <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/4-2-commute_times.csv")
citations <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/4-4-citations.csv")
academy_scores <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/4-1-academy_scores.csv")
```

## Example 2

- Do assessment scores differ across the three training academies (Meadowbrook, Coldharbor, and Northgate)?

```{r}
academy_scores |> anovas(outcome = score,
                         grouping = academy)
```

## Example 2

- Do assessment scores differ across the three training academies (Meadowbrook, Coldharbor, and Northgate)?

```{r}
#| fig-align: 'center'
#| echo: false
academy_scores |> anova_check(outcome = score, grouping = academy)
```


## Kruskal-Wallis

- If we break *either* ANOVA assumption, we should implement the nonparametric version, the Kruskal-Wallis.

    - The Kruskal-Wallis is an extension of the Wilcoxon rank sum (as ANOVA is an extension of the two-sample *t*-test).

- The Kruskal-Wallis test determines if $k$ independent samples come from populations with the same distribution. 

- Note: we can transform the data to make the variance "behave"

    - Box-Cox, or power, transformation


## Hypothesis Test for $M_1 = M_2 = ... = M_k$

- **Hypotheses**

    - $H_0: M_1 = ... = M_k$
    - $H_1:$ at least one $M_i$ is different

## Hypothesis Testing: Kruskal-Wallis

- **Test Statistic**

$$\chi^2_0 = \frac{12}{n(n+1)} \sum_{i=1}^k \frac{R_i^2}{n_i} - 3(n+1)  \sim \chi^2_{\text{df}}$$ 

- where
    
    - $R_i$ is the sum of the ranks for group $i$,
    - $n_i$ is the sample size for group $i$,
    - $n = \sum_{i=1}^k n_i$ = total sample size,
    - $k$ is the number of groups, and
    - $\text{df} = k-1$

## Statistical Inference on $k$ Medians (`R`)

- We will use the `kruskal_wallis()` function from `library(ssstats)` to perform the Kruskal-Wallis test.

```{r}
#| eval: false
dataset_name |> kruskal_wallis(outcome = continuous_variable,
                               grouping = grouping_variable,
                               alpha = specified_alpha)
```

## Example 2

```{r}
academy_scores |> kruskal_wallis(outcome = score,
                                 grouping = academy)
```
## Example 2

- **Hypotheses**

    - $H_0: M_{\text{M}} = M_{\text{C}} = M_{\text{N}}$
    - $H_1:$ at least one district's academy is different

- **Test Statistic and *p*-Value**

    - $\chi^2_0 = 9.801$
    - $p = 0.007$

- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Reject $H_0$. There is sufficient evidence to suggest that there is a difference in assessment scores between the three academies.

## Example 3

- Does the number of citations issued per officer differ across the four districts?

```{r}
#| fig-align: 'center'
citations |> anova_check(outcome = citations, grouping = district)
```

## Example 3

- Does the number of citations issued per officer differ across the four districts?

```{r}
#| fig-align: 'center'
citations |> filter(district == "Meadowbrook") |> one_qq(citations)
```

## Example 3

```{r}
citations |> kruskal_wallis(outcome = citations,
                            grouping = district)
```

## Example 3

- **Hypotheses**

    - $H_0: M_{\text{M}} = M_{\text{C}} = M_{\text{N}} = M_{\text{F}}$
    - $H_1:$ at least one district is different

- **Test Statistic and *p*-Value**

    - $\chi^2_0 = 43.119$
    - $p < 0.001$

- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Reject $H_0$. There is sufficient evidence to suggest that there is a difference in citations between the four districts.

## Wrap Up

- This lecture introduced the nonparametric alternative to one-way ANOVA.

    - Like ANOVA, the Kruskal-Wallis examines the effect of a single factor.
  
- The Kruskal-Wallis only allows us to determine if there is an overall, or omnibus, effect.

    - It does *not* tell us about pairwise differences.
    
- In the future:

    - Posthoc testing