---
title: "Posthoc Testing in One-Way Analyses"
execute:
  echo: true
  warning: false
  message: false
  error: true
format:
  revealjs:
    theme: simple
    css: style_lecture.css
    include-after-body: headers.html
    self-contained: true
    slide-number: true
    center: false
    width: 1600
    height: 900
    df-print: kable
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
    code-annotations: hover
editor: source
---

## Introduction

- Previous: one sample, two samples

- Now: more than two groups, continuous outcome

    - One-way ANOVA

    - Kruskal-Wallis

    - posthoc testing
    
- This lecture: posthoc testing

## `R` Setup 

```{r}
library(tidyverse)
library(ssstats)
commute_times <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/4-2-commute_times.csv")
citations <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/4-4-citations.csv")
academy_scores <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/4-1-academy_scores.csv")
response_times <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/4-3-response_times.csv")
```

## Posthoc Testing

- Recall our hypotheses in one-way ANOVA,

    - $H_0: \mu_1 = \mu_2 = ... = \mu_k$
    - $H_1:$ at least one $\mu_i$ is different
    
- The $F$ test does not tell us *which* mean is different... only that a difference exists.

- In theory, we could perform repeated $t$ tests to determine pairwise differences.
    - Recall that ANOVA is an extension of the $t$ test... or that the $t$ test is a special case of ANOVA.
    - However, this will increase the Type I error rate ($\alpha$).

## Posthoc Testing

- Recall that the Type I error rate, $\alpha$, is the probability of *incorrectly* rejecting $H_0$.

    - i.e., we are saying there is a difference between the means when there is actually *not* a difference.
    
- Suppose we are comparing 5 groups.

    - This is 10 pairwise comparisons!! 
        
        - 1-2, 1-3, 1-4, 1-5, 2-3, 2-4, 2-5, 3-4, 3-5, 4-5
        
    - If we perform repeated $t$ tests under $\alpha=0.05$, we are inflating the Type I error to 0.40! 😵

## Posthoc Testing

- When performing posthoc comparisons, we can choose one of two paths:

    - Control the Type I (familywise) error rate.
    - Do not control the Type I error rate.
    
- Note that controlling the Type I error rate is more conservative than when we do not control it.

    - "Conservative" = more difficult to reject.
    
- Generally, statisticians:

    - *do not* control the Type I error rate if examining the results of pilot/preliminary studies that are exploring for general relationships.
    
    - *do* control the Type I error rate if examining the results of confirmatory studies and are attempting to confirm relationships observed in pilot/preliminary studies.
    
## Posthoc Testing

- The posthoc tests we will learn:

    - Tukey's test
    
        - Performs all pairwise tests and controls the Type I error rate
        
    - Fisher's least significant difference 
    
        - Performs all pairwise tests but does not control the Type I error rate
        
- **Caution**: we should *only* perform posthoc tests if we have determined that a general difference exists!

    - i.e., when we reject when looking at either ANOVA or Kruskal-Wallis
    
## Posthoc Testing

- We will be performing *pairwise* comparisons.

- For each test performed, we are testing:

    - If using an ANOVA-based method:
    
        - $H_0: \ \mu_i = \mu_j$ 
        - $H_1: \ \mu_i \ne \mu_j$
        
    - If using a KW-based method:
    
        - $H_0: \ M_i = M_j$ 
        - $H_1: \ M_i \ne M_j$   
        
## Example 4

- Okonkwo again, "Dispatch keeps flagging Ashford as slow on 911 response, but Ashford's also our biggest district by square footage, so maybe that's just geography. Let's compare response times across districts."

- We will use the `response_times` dataset to answer the following question:

    - Do 911 response times differ across the five districts?
    
- First, let's check assumptions.    

## Example 4

```{r}
#| fig-align: 'center'
response_times |> anova_check(outcome = response_minutes,
                              grouping = district)
```

## Example 4

- Examining the variance assumption further,

```{r}
#| fig-align: 'center'
response_times |> multiple_variances(outcome = response_minutes,
                                     grouping = district)
```

## Example 4

- Now, looking at the omnibus test (ANOVA),

```{r}
response_times |> anovas(outcome = response_minutes,
                         grouping = district)
```        
        
## Posthoc Testing: Generic (R)

- We will use the `posthoc()` function from `library(ssstats)` to perform Tukey's posthoc test (resulting $p$ are adjusted for multiple comparisons).

```{r}
#| eval: false
dataset_name |> posthoc(outcome = continuous_variable, # <1>
                        grouping = grouping_variable, # <2>
                        adjust = TRUE or FALSE, # <3>
                        nonparametric = TRUE or FALSE, # <4>
                        alpha = alpha_level) # <5>
```
1. `outcome`: the same continuous variable used in your ANOVA or Kruskal-Wallis test.
2. `grouping`: the same grouping variable used in your ANOVA or Kruskal-Wallis test (must have at least 2 levels).
3. `adjust`: whether to correct for running multiple comparisons; default is `TRUE`. For ANOVA, `TRUE` runs Tukey's HSD and `FALSE` runs Fisher's LSD. For Kruskal-Wallis, `TRUE` applies a Bonferroni correction to Dunn's test and `FALSE` leaves it unadjusted.
4. `nonparametric`: `FALSE` for ANOVA-based methods (default); `TRUE` for Kruskal-Wallis based methods.
5. `alpha`: the significance level; default is `0.05`.        
    
## Posthoc Testing: Tukey's Test (ANOVA)

- Tukey's test allows us to do *all* pairwise comparisons while controlling $\alpha$.

- **Hypotheses**

    - $H_0: \ \mu_i = \mu_j$ 
    - $H_1: \ \mu_i \ne \mu_j$
    
- **Test Statistic**

$$
Q = \frac{|\bar{y}_i - \bar{y}_j|}{\sqrt{ \frac{\text{MS}_{\text{E}}}{2} \left( \frac{1}{n_i} + \frac{1}{n_j} \right) }}
$$


## Posthoc Testing: Tukey's (R)

- To use `posthoc()` for Tukey's test,

```{r}
#| eval: false
dataset_name |> posthoc(outcome = continuous_variable, # <1>
                        grouping = grouping_variable, # <2>
                        adjust = TRUE, # <3>
                        nonparametric = FALSE, # <4>
                        alpha = alpha_level) # <5>
```
1. `outcome`: the same continuous variable used in your ANOVA or Kruskal-Wallis test.
2. `grouping`: the same grouping variable used in your ANOVA or Kruskal-Wallis test (must have at least 2 levels).
3. `adjust`: whether to correct for running multiple comparisons; default is `TRUE`. For ANOVA, `TRUE` runs Tukey's HSD and `FALSE` runs Fisher's LSD. For Kruskal-Wallis, `TRUE` applies a Bonferroni correction to Dunn's test and `FALSE` leaves it unadjusted.
4. `nonparametric`: `FALSE` for ANOVA-based methods (default); `TRUE` for Kruskal-Wallis based methods.
5. `alpha`: the significance level; default is `0.05`.   

## Example 4

- Examining Tukey's for our example,

```{r}
response_times |> posthoc(outcome = response_minutes,
                          grouping = district,
                          adjust = TRUE,
                          nonparametric = FALSE)
```

## Posthoc Testing: Fisher's Test (ANOVA)

- Fisher's allows us to test *all* pairwise comparisons but *does not* control the $\alpha$.

- **Hypotheses:**
    - $H_0: \ \mu_i = \mu_j$
    - $H_1: \ \mu_i \ne \mu_j$

- **Test Statistic:**

$$
t = \frac{|\bar{y}_i - \bar{y}_j|}{\sqrt{ \text{MS}_{\text{E}} \left( \frac{1}{n_i} + \frac{1}{n_j} \right) }}
$$

## Posthoc Testing: Fisher's (R)

- To use `posthoc()` for Fisher's test,

```{r}
#| eval: false
dataset_name |> posthoc(outcome = continuous_variable, # <1>
                        grouping = grouping_variable, # <2>
                        adjust = FALSE, # <3>
                        nonparametric = FALSE, # <4>
                        alpha = alpha_level) # <5>
```
1. `outcome`: the same continuous variable used in your ANOVA or Kruskal-Wallis test.
2. `grouping`: the same grouping variable used in your ANOVA or Kruskal-Wallis test (must have at least 2 levels).
3. `adjust`: whether to correct for running multiple comparisons; default is `TRUE`. For ANOVA, `TRUE` runs Tukey's HSD and `FALSE` runs Fisher's LSD. For Kruskal-Wallis, `TRUE` applies a Bonferroni correction to Dunn's test and `FALSE` leaves it unadjusted.
4. `nonparametric`: `FALSE` for ANOVA-based methods (default); `TRUE` for Kruskal-Wallis based methods.
5. `alpha`: the significance level; default is `0.05`.   

## Example 4

- Examining Fisher's for our example,

```{r}
response_times |> posthoc(outcome = response_minutes,
                          grouping = district,
                          adjust = FALSE,
                          nonparametric = FALSE)
```

## Example 2

- Recall our second question: do assessment scores differ across the three training academies (Meadowbrook, Coldharbor, and Northgate)?

```{r}
#| fig-align: 'center'
#| echo: false
academy_scores |> anova_check(outcome = score, grouping = academy)
```

## Example 2

- Do assessment scores differ across the three training academies (Meadowbrook, Coldharbor, and Northgate)?

```{r}
academy_scores |> kruskal_wallis(outcome = score,
                                 grouping = academy)
```

## Posthoc Testing: Dunn's Test

- We can also perform posthoc testing in the Kruskal-Wallis setting using Dunn's test.

    - Rather than compare pairwise means, it compares pairwise average ranks.

- **Hypotheses:**
    - $H_0: \ M_{i} = M_{j}$
    - $H_1: \ M_{i} \ne M_{j}$

- **Test Statistic:**

$$
z_0 = \frac{|\bar{R}_i - \bar{R}_j|}{\sqrt{ \frac{n(n+1)}{12} \left( \frac{1}{n_i} + \frac{1}{n_j} \right) }}
$$

## Posthoc Testing: Adjusted Dunn's (R)

- To use `posthoc()` for Dunn's test adjusting for the Type I error rate,

```{r}
#| eval: false
dataset_name |> posthoc(outcome = continuous_variable, # <1>
                        grouping = grouping_variable, # <2>
                        adjust = TRUE, # <3>
                        nonparametric = TRUE, # <4>
                        alpha = alpha_level) # <5>
```
1. `outcome`: the same continuous variable used in your ANOVA or Kruskal-Wallis test.
2. `grouping`: the same grouping variable used in your ANOVA or Kruskal-Wallis test (must have at least 2 levels).
3. `adjust`: whether to correct for running multiple comparisons; default is `TRUE`. For ANOVA, `TRUE` runs Tukey's HSD and `FALSE` runs Fisher's LSD. For Kruskal-Wallis, `TRUE` applies a Bonferroni correction to Dunn's test and `FALSE` leaves it unadjusted.
4. `nonparametric`: `FALSE` for ANOVA-based methods (default); `TRUE` for Kruskal-Wallis based methods.
5. `alpha`: the significance level; default is `0.05`.  

## Example 2

- Examining adjusted Dunn's for our example,

```{r}
academy_scores |> posthoc(outcome = score,
                          grouping = academy,
                          adjust = TRUE,
                          nonparametric = TRUE)
```

## Posthoc Testing: Unadjusted Dunn's (R)

- To use `posthoc()` for Dunn's test without adjusting for the Type I error rate,

```{r}
#| eval: false
dataset_name |> posthoc(outcome = continuous_variable, # <1>
                        grouping = grouping_variable, # <2>
                        adjust = FALSE, # <3>
                        nonparametric = TRUE, # <4>
                        alpha = alpha_level) # <5>
```
1. `outcome`: the same continuous variable used in your ANOVA or Kruskal-Wallis test.
2. `grouping`: the same grouping variable used in your ANOVA or Kruskal-Wallis test (must have at least 2 levels).
3. `adjust`: whether to correct for running multiple comparisons; default is `TRUE`. For ANOVA, `TRUE` runs Tukey's HSD and `FALSE` runs Fisher's LSD. For Kruskal-Wallis, `TRUE` applies a Bonferroni correction to Dunn's test and `FALSE` leaves it unadjusted.
4. `nonparametric`: `FALSE` for ANOVA-based methods (default); `TRUE` for Kruskal-Wallis based methods.
5. `alpha`: the significance level; default is `0.05`.  

## Example 2

- Examining unadjusted Dunn's for our example,

```{r}
academy_scores |> posthoc(outcome = score,
                          grouping = academy,
                          adjust = FALSE,
                          nonparametric = TRUE)
```

## Example 1

- Recall our first question: does commute time differ across the four districts?

```{r}
commute_times |> anovas(outcome = commute_minutes,
                        grouping = district)
```


## Example 1

- Halt!

- $p = 0.507$ $\to$ FTR $\to$ we did *not* detect a difference in the means.

- **We do not perform posthoc testing when ANOVA or Kruskal-Wallis did *not* show an overall difference.**

- Remember that we said ANOVA and Kruskal-Wallis were omnibus tests.

    - If the omnibus test does not suggest a difference exists, we do not further poke around.

## Wrap Up

- This lecture introduced posthoc testing, the follow-up to ANOVA and Kruskal-Wallis.

    - If ANOVA $\to$ Tukey's (adjusted) or Fisher's (unadjusted)
    
    - If Kruskal-Wallis $\to$ adjusted Dunn's or unadjusted Dunn's
    
- We only perform posthoc testing when the omnibus test (ANOVA or Kruskal-Wallis) has determined there is an overall difference in the means or medians.    

- We also discussed the familywise error rate (FWE), $\alpha$.

    - When we *adjust* to preserve the FWE, we maintain $\alpha = 0.05$, regardless of the number of tests performed.
    
    - When we *do not adjust* to preserve the FWE, we allow $\alpha$ to inflate; the inflation depends on number of pairwise comparisons.
  
