---
title: "Two-Way ANOVA"
execute:
  echo: true
  warning: false
  message: false
  error: true
format:
  revealjs:
    theme: simple
    css: style_lecture.css
    include-after-body: headers.html
    self-contained: true
    slide-number: true
    center: false
    width: 1600
    height: 900
    df-print: kable
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
    code-annotations: hover
editor: source
---

## Introduction

- Recall that ANOVA allows us to compare the means of three or more groups.

- In one-way ANOVA, we are only considering one factor (grouping variable). 

- In this module, we discuss two-way ANOVA, which allows us to consider a second factor (grouping variable). 

## `R` Setup 

```{r}
library(tidyverse)
library(ssstats)
exam_scores <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/5-1-exam_scores.csv")
fatigue_scores <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/5-2-fatigue_scores.csv")
```

## Two-Way ANOVA

- Recall in one-way ANOVA, we partitioned SS<sub>Tot</sub> into SS<sub>Trt</sub> and SS<sub>E</sub>.

- Now that multiple factors are involved, we now partition the SS<sub>Trt</sub> into the different factors.

    - Here, that is SS<sub>A</sub>, SS<sub>B</sub>, and SS<sub>AB</sub>.
    
- Recall that SS<sub>E</sub> is the "catch all" for unexplained variance. 

    - When we add factors to our model, we are moving part of the SS<sub>E</sub> into the SS<sub>Trt</sub>.
    
## Two-Way ANOVA

- Terminology in two-way ANOVA. 

- Factor A has $a$ levels.
    
    - Donut flavor: glazed, chocolate, maple

- Factor B has $b$ levels.

    - Species: wolf, rabbit, elephant

- There are $ab$ "treatment groups."

    - Every observation will belong to one level of $a$ and one level of $b$.

## Interaction Effects

- Suppose Okonkwo is testing a new donut recipe for the break room. He has officers taste-test it and rate how much they liked it.

- The outcome is the raters' taste score from 1–10.

- Then, he's varying two things at once:

    - Donut flavor: which flavor was baked (Glazed, Chocolate, Maple)
    
    - Species: what kind of animal is doing the tasting (Wolf, Rabbit, Elephant)
    
## Interaction Effects

- This design allows us to check for an **interaction effect**, implying the two factors don't act independently on the outcome.

    — i.e., the effect of one factor on the outcome depends on the level of the other factor.

- In Okonkwo's donut test, an interaction implies both:

    - The relationship between rating and donut flavor *depends on* species.
    - The relationship between rating and species *depends on* donut flavor.

- Remember, these are two descriptions of the *same* interaction, not two separate effects.

## Two-Way ANOVA: Table

- The table we are working towards:

<table class="anova-table">
  <thead>
    <tr>
      <th>Source</th>
      <th>Sum of Squares</th>
      <th>df</th>
      <th>Mean Square</th>
      <th><em>F</em></th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td class="treatment">A</td>
      <td class="summary">SS<sub>A</sub></td>
      <td class="summary">df<sub>A</sub></td>
      <td class="summary">MS<sub>A</sub></td>
      <td class="summary"><em>F</em><sub>A</sub></td>
    </tr>
    <tr>
      <td class="treatment">B</td>
      <td class="summary">SS<sub>B</sub></td>
      <td class="summary">df<sub>B</sub></td>
      <td class="summary">MS<sub>B</sub></td>
      <td class="summary"><em>F</em><sub>B</sub></td>
    </tr>
    <tr>
      <td class="treatment">AB</td>
      <td class="summary">SS<sub>AB</sub></td>
      <td class="summary">df<sub>AB</sub></td>
      <td class="summary">MS<sub>AB</sub></td>
      <td class="summary"><em>F</em><sub>AB</sub></td>
    </tr>
    <tr>
      <td class="treatment">Error</td>
      <td class="summary">SS<sub>E</sub></td>
      <td class="summary">df<sub>E</sub></td>
      <td class="summary">MS<sub>E</sub></td>
      <td></td>
    </tr>
    <tr>
      <td class="treatment">Total</td>
      <td class="summary">SS<sub>Tot</sub></td>
      <td class="summary">df<sub>Tot</sub></td>
      <td></td>
      <td></td>
    </tr>
  </tbody>
</table> 

## Two-Way ANOVA: Computation

- Let there be $a$ levels of factor A and $b$ levels of factor B. 

    - $y_{ijk}$ is the observation on the $k^{th}$ experimental unit receiving the $i^{th}$ level of factor A and the $j^{th}$ level of factor B 
    - $y_{i.}$ is the sum for all observations at the $i^{th}$ level of factor A,
    - $y_{.j}$ is the sum for all observations at the $j^{th}$ level of factor B,
    - $y_{ij}$ is the sum for observations at the $i^{th}$ level of factor A and the $j^{th}$ level of factor B,
    - $y_{..}$ is the sum of all observations, 
    - $(y^2)_{..}$ is the sum of the squared observations, 
    - $n$ is the number in each group (of $a \times b$ treatments)

## Two-Way ANOVA: Computation

- To find the SS and df:
$$
\begin{align*}
\text{SS}_{\text{A}} &= \frac{\sum_i y_{i.}^2}{bn}  - \frac{(y_{..})^2}{abn} & \text{df}_{\text{A}} &= a-1 \\
\text{SS}_{\text{B}} &= \frac{\sum_j y_{.j}^2}{an} - \frac{(y_{..})^2}{abn} & \text{df}_{\text{B}} &= b-1 \\
\text{SS}_{\text{AB}}&= \frac{\sum_{ij} y_{ij}^2}{n} - \frac{(y_{..})^2}{abn} - \text{SS}_{\text{A}} - \text{SS}_{\text{B}} & \text{df}_{\text{AB}} &= (a-1)(b-1) \\
\text{SS}_{\text{E}} &= \text{SS}_{\text{Tot}} - \text{SS}_{\text{A}} - \text{SS}_{\text{B}} - \text{SS}_{\text{AB}} & \text{df}_{\text{E}} &= ab(n-1)  \\
\text{SS}_{\text{Tot}} &= (y^2)_{..} - \frac{(y_{..})^2}{abn} & \text{df}_{\text{Tot}} &= abn-1
\end{align*}  
$$  

## Two-Way ANOVA: Computation

- To find the MS:

$$\text{MS}_{\text{X}} = \frac{\text{SS}_{\text{X}}}{\text{df}_{\text{X}}}$$

- To find the test statistics:

$$\text{F}_{\text{X}} = \frac{\text{MS}_{\text{X}}}{\text{MS}_{\text{E}}}$$

## Two-Way ANOVA (`R`)

- To get the two-way ANOVA table in R,

```{r}
#| eval: false
dataset_name |> anovas(outcome = response_variable, # <1>
                       A = factor_A, # <2>
                       B = factor_B, # <3>
                       interaction = TRUE or FALSE # <4>
  )
```
1. `outcome`: the continuous variable being compared across groups.
3. `A`: the first grouping variable for a two-way ANOVA. Supplying `A` overrides `grouping` — `anovas()` treats this as a two-way analysis.
4. `B`: the second grouping variable for a two-way ANOVA. Supplying `B` requires `A` to also be specified.
5. `interaction`: whether to fit the interaction term along with both main effects; default is `TRUE`. `TRUE` fits `outcome ~ A * B` and reports the interaction test only; `FALSE` fits `outcome ~ A + B` (main effects only, no interaction term) and reports both main-effect tests instead.

## Hypothesis Test for an Interaction

- **Hypotheses**

    - $H_0:$ there is not an interaction between [factor A] and [factor B]
    - $H_1:$ there is an interaction between [factor A] and [factor B]

- **Test Statistic and *p*-Value**

    - $F_{\text{AB}}$ 
    - $p = P[F_{\text{df}_{\text{AB}}, \text{df}_{\text{E}}} \ge F_{\text{AB}}]$

## Hypothesis Test for an Interaction (`R`)

```{r}
#| eval: false
dataset_name |> anovas(outcome = response_variable,
                       A = factor_A,
                       B = factor_B,
                       interaction = TRUE,
                       alpha = alpha_level)
```

## Example 1

- Okonkwo pulls up a report on your screen: "Wellness office wants to know: does night shift wear people out the same way across every unit?"

- We will use the `fatigue_scores` dataset to answer the following question:

    - Does the effect of shift (Day vs. Night) on fatigue depend on which unit an officer works in?
    
## Example 1

- Exploring the levels in shift and unit,

```{r}
fatigue_scores |> n_pct(shift, unit)
```

## Example 1

- Does the effect of shift on fatigue depend on which unit an officer works in?

```{r}
fatigue_scores |> anovas(outcome = fatigue_score,
                         A = shift,
                         B = unit)
```

## Example 1

- **Hypotheses**

    - $H_0:$ there is not an interaction between shift and unit
    - $H_1:$ there is an interaction between shift and unit

- **Test Statistic and *p*-Value**

    - $F_0 = 3.753$
    - $p = 0.025$

- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Reject $H_0$. There is sufficient evidence to suggest that there is an interaction between shift and unit.    

## Example 2

- Okonkwo again: "Training division ran their numbers and thinks the new Enhanced program is worth the extra budget, but I want a second opinion. Does the training method work the same for everybody?"

- We will use the `exam_scores` dataset to answer the following question:

    - Does the effect of training method depend on experience level?
    
## Example 2

- Exploring the levels in training method and experience level,

```{r}
exam_scores |> n_pct(training_method, experience_level)
```

## Example 2

- Does the effect of training method depend on experience level?

```{r}
exam_scores |> anovas(outcome = exam_score,
                      A = training_method,
                      B = experience_level)
```

## Example 2

- **Hypotheses**

    - $H_0:$ there is not an interaction between training method and experience level
    - $H_1:$ there is an interaction between training method and experience level

- **Test Statistic and *p*-Value**

    - $F_0 = 0.027$
    - $p = 0.973$

- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Fail to reject $H_0$. There is not sufficient evidence to suggest that there is an interaction between training method and experience level.    

## Two-Way ANOVA: Testing Main Effects

- What if the interaction *is not* significant?
    - We want to *remove* the interaction and examine the *main effects*.
    - That is, we want to look at the individual factors.
    
- We will reconstruct the ANOVA table. If we were doing this by hand:

    1. Rewrite ANOVA table without interaction; do not change SS, df, and MS for the main effects and total.
    2. Update the error term: add the SS<sub>AB</sub> to SS<sub>E</sub> and df<sub>AB</sub> to df<sub>E</sub>.
    3. Recalculate MS<sub>E</sub>.
    4. Recalculate *F*<sub>A</sub> and *F*<sub>B</sub>, then find the corresponding *p*-values.

## Hypothesis Test for Main Effects   

- **Hypotheses**

    - $H_0:$ $\mu_1$ = $\mu_2$ = ... = ($\mu_a$ or $\mu_b$)
    - $H_1:$ At least one $\mu_i$ is different.

- **Test Statistic and *p*-Value**

    - $F_{\text{X}}$ 
    - $p = P[F_{\text{df}_{\text{X}}, \text{df}_{\text{E}}} \ge F_{\text{X}}]$

## Hypothesis Test for Main Effects (`R`)

```{r}
#| eval: false
dataset_name |> anovas(outcome = response_variable,
                       A = factor_A,
                       B = factor_B,
                       interaction = FALSE,
                       alpha = alpha_level)
```

## Example 2

- Are either training method or experience level related to exam scores?

```{r}
exam_scores |> anovas(outcome = exam_score,
                      A = training_method,
                      B = experience_level,
                      interaction = FALSE)
```

## Example 2

- **Hypotheses**

    - $H_0: \mu_{\text{E}} = \mu_{\text{P}} = \mu_{\text{S}}$
    - $H_1:$ at least one training method is different

- **Test Statistic and *p*-Value**

    - $F_0 = 7.669$
    - $p < 0.001$

- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Reject $H_0$. There is sufficient evidence to suggest that there is a difference between the training methods.

## Example 2

- **Hypotheses**

    - $H_0: \mu_{\text{R}} = \mu_{\text{V}}$
    - $H_1: \mu_{\text{R}} \ne \mu_{\text{V}}$

- **Test Statistic and *p*-Value**

    - $F_0 = 1.018$
    - $p = 0.314$

- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$

- **Conclusion and Interpretation**

    - Fail to reject $H_0$. There is not sufficient evidence to suggest a difference between rookies and veterans.

## Wrap Up

- In this lecture, we have covered two-way ANOVA.

    - Test for interaction.
    
    - Test for main effects.
    
- In our following lectures, we will discuss next steps after initial hypothesis testing in two-way ANOVA.    
    
    - posthoc testing
    
    - profile plots