---
title: "Posthoc Testing in Two-Way ANOVA"
execute:
  echo: true
  warning: false
  message: false
  error: true
format:
  revealjs:
    theme: simple
    css: style_lecture.css
    include-after-body: headers.html
    self-contained: true
    slide-number: true
    center: false
    width: 1600
    height: 900
    df-print: kable
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
    code-annotations: hover
editor: source
---

## Introduction

- Recall that ANOVA allows us to compare the means of three or more groups.

- In one-way ANOVA, we are only considering one factor (grouping variable). 

- In this module, we discuss two-way ANOVA, which allows us to consider a second factor (grouping variable). 

- We now will discuss performing posthoc tests in the two-way ANOVA setting.

## `R` Setup 

```{r}
library(tidyverse)
library(ssstats)
exam_scores <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/5-1-exam_scores.csv")
fatigue_scores <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/5-2-fatigue_scores.csv")
```

## Posthoc Testing

- We can follow up two-way ANOVA tests with posthoc tests.

- Recall our hypotheses for main effects,

    - $H_0: \mu_1 = \mu_2 = ... = \mu_a$
    - $H_1:$ at least one $\mu_i$ is different
    
- We will again turn to posthoc tests to determine pairwise differences.

## Posthoc Testing (`R`)

- We will again use the `posthoc()` function,

```{r}
#| eval: false
dataset_name |> posthoc(outcome = response_variable,
                        A = factor_A,
                        B = factor_B,
                        interaction = TRUE or FALSE,
                        adjust = TRUE or FALSE)
```

## Example 1

- Recall the two-way ANOVA results for the `fatigue_scores` data,

- Does the effect of shift on fatigue depend on which unit an officer works in?

```{r}
fatigue_scores |> anovas(outcome = fatigue_score,
                         A = shift,
                         B = unit)
```

## Example 1

- We can look at the posthoc test of the interaction term.

- Treating it as exploratory,

```{r}
fatigue_scores |> posthoc(outcome = fatigue_score,
                       A = shift,
                       B = unit,
                       interaction = TRUE,
                       adjust = FALSE)
```

## Example 2

- Recall the two-way ANOVA results for the `exam_scores` data,

- Does the effect of training method depend on experience level?

```{r}
exam_scores |> anovas(outcome = exam_score,
                      A = training_method,
                      B = experience_level,
                      interaction = TRUE)
```

## Example 2

- Recall the two-way ANOVA results for the `exam_scores` data,

- Are there main effects of training method or experience level?

```{r}
exam_scores |> anovas(outcome = exam_score,
                      A = training_method,
                      B = experience_level,
                      interaction = FALSE)
```

## Example 2

- We can look at the posthoc tests of the main effects.

- Treating it as confirmatory,

```{r}
fatigue_scores |> posthoc(outcome = fatigue_score,
                       A = shift,
                       B = unit,
                       interaction = TRUE,
                       adjust = TRUE)
```

## Wrap Up

- In this lecture, we have covered posthoc testing in two-way ANOVA.

    - Interactions.
    
    - Main effects.
    
- In our following lectures, we will discuss next steps.    
    
    - profile plots

