---
title: "Profile Plots in Two-Way ANOVA"
execute:
  echo: true
  warning: false
  message: false
  error: true
format:
  revealjs:
    theme: simple
    css: style_lecture.css
    include-after-body: headers.html
    self-contained: true
    slide-number: true
    center: false
    width: 1600
    height: 900
    df-print: kable
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
    code-annotations: hover
editor: source
---

## Introduction

- Recall that ANOVA allows us to compare the means of three or more groups.

- In one-way ANOVA, we are only considering one factor (grouping variable). 

- In this module, we discuss two-way ANOVA, which allows us to consider a second factor (grouping variable). 

- In this lecture, we discuss profile plots, a way to visualize treatment group means.

## `R` Setup 

```{r}
library(tidyverse)
library(ssstats)
exam_scores <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/5-1-exam_scores.csv")
fatigue_scores <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/5-2-fatigue_scores.csv")
```

## Two-Way ANOVA: Profile Plots

- We can construct profile plots  to display treatment group means.

    - *y*-axis: always the average outcome
    
    - *x*-axis: either factor A or B
    
    - "lines" on the plot: the factor that was not selected for the *x*-axis

- Note that this is just a graph of the means! 

    - It's valid to construct a profile plot regardless of significance.
    
## Two-Way ANOVA: Profile Plots (R)

- We will use the `profile_plot()` function from `library(ssstats)` to construct basic profile plots.

```{r}
#| eval: false
dataset_name |> profile_plot(continuous = continuous_variable,
                              xaxis = variable_on_x,
                              lines = variable_for_lines)
```    

## Example 1

- Recall the two-way ANOVA results for the `fatigue_scores` data,

- Does the effect of shift on fatigue depend on which unit an officer works in?

```{r}
fatigue_scores |> anovas(outcome = fatigue_score,
                         A = shift,
                         B = unit)
```

## Example 1

- Let's remind ourselves what's "in" the variables:

```{r}
fatigue_scores |> n_pct(shift, unit)
```

- Neither of the factors has a natural "order"

- Let's have `unit` on the *x*-axis and define the lines by `shift`.

## Example 1

- Constructing the profile plot,

```{r}
#| fig-align: 'center'
fatigue_scores |> profile_plot(outcome = fatigue_score,
                              xaxis = unit,
                              lines = shift)
```  










## Example 2

- Recall the two-way ANOVA results for the `exam_scores` data,

- Does the effect of training method depend on experience level?

```{r}
exam_scores |> anovas(outcome = exam_score,
                      A = training_method,
                      B = experience_level,
                      interaction = TRUE)
```

## Example 2

- Recall the two-way ANOVA results for the `exam_scores` data,

- Are there main effects of training method or experience level?

```{r}
exam_scores |> anovas(outcome = exam_score,
                      A = training_method,
                      B = experience_level,
                      interaction = FALSE)
```

## Example 2

- Let's remind ourselves what's "in" the variables:

```{r}
exam_scores |> n_pct(training_method, experience_level)
```

- Neither of the factors has a natural "order"

- Let's have `training_method` on the *x*-axis and define the lines by `experience_level`.

## Example 2

- Constructing the profile plot,

```{r}
#| fig-align: 'center'
exam_scores |> profile_plot(outcome = exam_score,
                            xaxis = training_method,
                            lines = experience_level)
```  

## Wrap Up

- In this lecture, we have covered profile plots in two-way ANOVA.

- This lecture concludes the two-way ANOVA module.

- Next module: categorical analysis.