---
title: "Scatterplots and Correlation"
execute:
  echo: true
  warning: false
  message: false
  error: true
format:
  revealjs:
    theme: simple
    css: style_lecture.css
    include-after-body: headers.html
    self-contained: true
    slide-number: true
    center: false
    width: 1600
    height: 900
    df-print: kable
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
    code-annotations: hover
editor: source
---

## Introduction

- In this module, we begin discussing regression analysis.

- Regression is a very, very wide area. 

    - We are barely scratching the surface.
    
- SDS II will deepen your understanding on statistical modeling techniques. 

## `R` Set Up

```{r}
library(tidyverse)
library(ssstats)
training <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/7-1-training.csv")
wellness <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/7-2-wellness.csv")
reports <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/7-3-reports.csv")
break_room <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/7-4-break_room.csv")
```

## Scatterplots

- A **scatterplot** is a graph that shows the relationship between two quantitative variables measured on the same individual.

- Every individual in the dataset is represented by a point on the scatterplot.

- Note! The explanatory variable is on the $x$-axis and the response variable is on the $y$-axis.

- Like we have seen before, plotting the data is a first step in identifying issues.


## Scatterplots

- **Positive relationship:** As $x$ increases, $y$ also increases.

```{r}
#| echo: false
#| fig-align: 'center'
library(tidyverse)
set.seed(123)
dat <- tibble(x = 1:20, y = 2*x + rnorm(20, 0, 4))

ggplot(dat, aes(x, y)) +
  geom_point() +
  theme_minimal(base_size = 16) +
  labs(x = "Explanatory variable (x)",
       y = "Response variable (y)")
```

## Scatterplots

- **Negative relationship:** As $x$ increases, $y$ decreases.

```{r}
#| echo: false
#| fig-align: 'center'
set.seed(123)
dat <- tibble(x = 1:20, y = 40 - 2*x + rnorm(20, 0, 4))

ggplot(dat, aes(x, y)) +
  geom_point() +
  theme_minimal(base_size = 16) +
  labs(x = "Explanatory variable (x)",
       y = "Response variable (y)")
```


## Scatterplots

- **No relationship:** As $x$ increases, we cannot predict what happens to $y$.

```{r}
#| echo: false
#| fig-align: 'center'
set.seed(123)

dat <- tibble(
  x = runif(20, 0, 20),
  y = runif(20, 0, 40)
)

ggplot(dat, aes(x, y)) +
  geom_point() +
  theme_minimal(base_size = 16) +
  labs(x = "Explanatory variable (x)",
       y = "Response variable (y)")
```

## Scatterplots

- Hmmmm...

```{r}
#| echo: false
#| fig-align: 'center'
set.seed(123)

dat <- tibble(
  x = seq(-5, 5, length.out = 20),
  y = x^2 + rnorm(20, 0, 2)
)

ggplot(dat, aes(x, y)) +
  geom_point() +
  theme_minimal(base_size = 16) +
  labs(x = "Explanatory variable (x)",
       y = "Response variable (y)")
```

## Scatterplots

- Hmmmm...

```{r}
#| echo: false
#| fig-align: 'center'
set.seed(123)

dat <- tibble(
  x = 1:20,
  y = 2*x + rnorm(20, 0, 4)
) |>
  add_row(
    x = c(5, 17),
    y = c(45, 5)
  )

ggplot(dat, aes(x, y)) +
  geom_point() +
  theme_minimal(base_size = 16) +
  labs(x = "Explanatory variable (x)",
       y = "Response variable (y)")
```

## Scatterplot (`R`)

- To construct a scatterplot in `R`, the base code uses `ggplot()` from the `ggplot2` package (called in with `tidyverse`).

```{r}
#| eval: false
dataset_name |> ggplot(aes(x = x_variable, y = y_variable)) +
  geom_point() +
  theme_minimal() +
  labs(x = "Explanatory variable (x)",
       y = "Response variable (y)")
```

## Example 1

- Okonkwo stops by your desk: "Training division keeps encouraging recruits to spend more time on practice exercises. Before we build extra practice time into the schedule, I want to know if recruits who practice more tend to score higher?"

- We will examine the following question using the `training` dataset:

    - Is there a relationship between weekly practice hours ($x$) and exam score ($y$)?
    
```{r}
#| echo: false
head(training, n = 3)
```

## Example 1

- Constructing the scatterplot,

```{r}
#| fig-align: 'center'
#| echo: false
training |> ggplot(aes(x = practice_hours, y = exam_score)) +
  geom_point() +
  theme_minimal(base_size = 16) +
  labs(x = "Practice Time (hrs)",
       y = "Exam Score")
```


## Example 2

- Okonkwo raises an eyebrow: "Someone in records claims our taller officers write more accurate incident reports. ... That can't be true. Is there actually a relationship?"

- We will examine the following question using the `reports` dataset:

    - Is there a relationship between officer height ($x$) and report accuracy ($y$)?


```{r}
#| echo: false
head(reports, n = 3)
```

## Example 2

- Constructing the scatterplot,

```{r}
#| fig-align: 'center'
#| echo: false
reports |> ggplot(aes(x = height_cm, y = report_accuracy)) +
  geom_point() +
  theme_minimal(base_size = 16) +
  labs(x = "Height (cm)",
       y = "Accuracy (%)")
```

## Correlation

- **Correlation:** A measure of the strength and direction of the linear relationship between two quantitative variables.

    - $\rho$ represents the population correlation coefficient.
    
    - $r$ represents the sample correlation coefficient.

- Note!! Correlation is bounded to [-1, 1].

    - $r=-1$ represents perfect negative correlation.
    - $r=1$ represents perfect positive correlation.
    - $r=0$ represents no correlation.
    
- The closer |$r$| is to 1, the *stronger* the relationship between $x$ and $y$.    

## Pearson's Correlation Coefficient

- **Pearson's correlation coefficient**:

$$r = \frac{\sum_{i=1}^n \left( \frac{x_i - \bar{x}}{s_x} \right)\left( \frac{y_i - \bar{y}}{s_y} \right)}{n-1},$$

- where

    - $x_i$ and $y_i$ make up the the $i$^th^ observation 
    - $\bar{x}$ and $\bar{y}$ are the sample means
    - $s_x$ and $s_y$ are the sample standard deviations
    - $n$ is the total sample size
    
## Pearson's Correlation Coefficient

- We can also determine if a **linear** relationship exists.

- **Hypotheses:**

    - $H_0: \rho = 0$ 
    - $H_1: \rho \ne 0$
    
- **Test Statistic:**

$$t_0 = \frac{r \sqrt{n-2}}{\sqrt{1-r^2}} \sim t_{n-2}$$
    
## Pearson's Correlation (`R`)

- We will use the `correlation()` function from `library(ssstats)`.

- Generic syntax:

```{r}
#| eval: false
dataset_name |> correlation(x = 1st_continuous_variable, # <1>
                            y = 2nd_continuous_variable) # <2>
```
1. the first continuous variable in the correlation.
2. the second continuous variable in the correlation.

## Example 1

- Finding the correlation for the `training` dataset,

    - Is there a relationship between weekly practice hours ($x$) and exam score ($y$)?

```{r}
training |> correlation(x = practice_hours, y = exam_score)
```

- Note that the choice (order) of $x$ and $y$ are arbitrary.

```{r}
training |> correlation(y = practice_hours, x = exam_score)
```

## Example 1

- **Hypotheses:**

    - $H_0: \rho = 0$ 
    - $H_1: \rho \ne 0$
    
- **Test Statistic and $p$-Value:**

    - $t_0 = 8.837$
    - $p < 0.001$
    
- **Rejection Region:**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$
    
- **Conclusion and Interpretation:**    

    - Reject $H_0$. There is sufficient evidence to suggest that there is a linear relationship between the practice time and exam score.

## Example 2

- Finding the correlation for the `reports` dataset,

    - Is there a relationship between officer height ($x$) and report accuracy ($y$)?

```{r}
reports |> correlation(height_cm, report_accuracy)
```

## Example 2

- **Hypotheses:**

    - $H_0: \rho = 0$ 
    - $H_1: \rho \ne 0$
    
- **Test Statistic and $p$-Value:**

    - $t_0 = 0.285$
    - $p = 0.777$
    
- **Rejection Region:**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$
    
- **Conclusion and Interpretation:**    

    - Fail to reject $H_0$. There is not sufficient evidence to suggest that there is a linear relationship between height and report accuracy.

## Assumptions on Pearson's Correlation (`R`)

- An assumption for Pearson's correlation is that both $x$ and $y$ are normally distributed. 

- We will check this using a q-q plot as we have in the past.

- We will use the `correlation()` function from `library(ssstats)`.

- Generic syntax:

```{r}
#| eval: false
dataset_name |> correlation_qq(x = 1st_continuous_variable, # <1>
                               y = 2nd_continuous_variable) # <2>
```
1. the first continuous variable in the correlation.
2. the second continuous variable in the correlation.

## Example 1

- Checking the assumption,

```{r}
#| fig-align: 'center'
training |> correlation_qq(practice_hours, exam_score)
```

## Example 2

- Checking the assumption,

```{r}
#| fig-align: 'center'
reports |> correlation_qq(height_cm, report_accuracy)
```

## Spearman's Correlation Coefficient

- Spearman's correlation coefficient is a nonparametric alternative to Pearson's correlation.
    
    - Spearman's correlation only assumes that the data is ordinal. 
    
    - It *does not* work for nominal data.

- To compute Spearman's, the following algorithm is used:

    1. Rank the $x$ and, separately, the $y$ values. 
    
        - ($x$, $y$) $\to$ ($R_x$, $R_y$)
    
    2. Find Pearson's correlation for the ranked data. 
    
- Spearman's correlation is interpreted the same as Pearson's correlation.   

## Pearson's Correlation Coefficient

- We can also determine if a **monotonic** relationship exists.

- **Hypotheses:**

    - $H_0: \rho_S = 0$ 
    - $H_1: \rho_S \ne 0$
    
- **Test Statistic:**

$$S = \frac{n(n^2 - 1)}{6}(1 - r_s)$$

## Spearman's Correlation (`R`)

- We will again use the `correlation()` function from `library(ssstats)`.

- Generic syntax:

```{r}
#| eval: false
dataset_name |> correlation(x = 1st_continuous_variable, # <1>
                            y = 2nd_continuous_variable, # <2>
                            method = "spearman") # <3> 
```
1. the first continuous variable in the correlation.
2. the second continuous variable in the correlation.
3. request Spearman's correlation instead of Pearson's.

## Example 2

- Finding the Spearman's correlation for the `reports` dataset,

    - Is there a relationship between officer height ($x$) and report accuracy ($y$)?

```{r}
reports |> correlation(height_cm, report_accuracy,
                       method = "spearman")
```

## Example 2

- **Hypotheses:**

    - $H_0: \rho_S = 0$ 
    - $H_1: \rho_S \ne 0$
    
- **Test Statistic and $p$-Value:**

    - $S_0 = 10396.93$
    - $p = 0.880$
    
- **Rejection Region:**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$
    
- **Conclusion and Interpretation:**    

    - Fail to reject $H_0$. There is not sufficient evidence to suggest that there is a linear relationship between height and report accuracy.


## Wrap Up

- guessthecorrelation.com!

- In the next lectures, we will learn how to quantify the relationship between $x$ and $y$ using regression analysis.

