---
title: "General Linear Models:<br>Estimation and Interpretation"
execute:
  echo: true
  warning: false
  message: false
  error: true
format:
  revealjs:
    theme: simple
    css: style_lecture.css
    include-after-body: headers.html
    self-contained: true
    slide-number: true
    center: false
    width: 1600
    height: 900
    df-print: kable
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
    code-annotations: hover
editor: source
---

## Introduction

- We have discussed quantifying the relationship between two continuous variables.

- Recall that the correlation describes the strength and the direction of the relationship.

    - Pearson's correlation: describes the linear relationship; assumes normality of both variables.
    
    - Spearman's correlation: describes the monotone relationship; assumes both variables are at least ordinal.

- We now focus on linear regression, an alternative way to quantify the relationship.

## `R` Set Up

```{r}
library(tidyverse)
library(ssstats)
library(broom)
training <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/7-1-training.csv")
wellness <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/7-2-wellness.csv")
reports <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/7-3-reports.csv")
break_room <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/7-4-break_room.csv")
bakery_sales <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/7-5-bakery_sales.csv")
cafe_sales <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/7-6-cafe_sales.csv")
```

## Introduction to Regression

- We are often interested in how two variables are related.

- We use regression analysis to examine this relationship.
    
    - One variable predicts the other.

- We will create what is called:

    - regression "line"
    - regression model
    - line of best fit
    - trend line

## Introduction to Regression

- Consider the general form of regression:

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + ... + \beta_k x_k + \varepsilon$$


- **Outcome** or response variable ($y$): the variable we are interested in predicting or modeling.

- **Predictor** or explanatory variable ($x_i$): a variable explaining or predicting the outcome.
    
    
## Introduction to Regression

- The regression equivalent to correlation is *simple* linear regression,

$$y = \beta_0 + \beta_1 x + \varepsilon,$$

- where
    
    - $\beta_0$ is the $y$-intercept,
    
    - $\beta_1$ is the *slope* describing the relationship between $x$ and $y$,
    
    - $\varepsilon$ is the *error*, or *residual*.
    
## Regression: Estimation

- When we only have one predictor in the model, we can easily estimate $\beta_0$ and $\beta_1$:

$$
\begin{align*}
\hat{\beta}_1 &= \frac{\sum_{i=1}^n (x_i - \bar{x})(y_i - \bar{y})}{\sum_{i=1}^n (x_i - \bar{x})^2} \\
    & \\
    \hat{\beta}_0 &= \bar{y} - \hat{\beta}_1 \bar{x}
\end{align*}
$$

## Regression: Estimation (`R`)

- We will use `glm()` to construct our linear models.

- For simple linear regression, we will first construct the model,

```{r}
#| eval: false
m <- glm(y ~ x,
         data = dataset_name)
```

- then we can look at the results using `coefficients()`.

```{r}
#| eval: false
m |> coefficients()
```

## Example 1

- Recall the `training` dataset,

    - Is there a relationship between weekly practice hours ($x$) and exam score ($y$)?
    
```{r}
#| echo: false
head(training, n = 3)
```

## Example 1

- Is there a relationship between weekly practice hours ($x$) and exam score ($y$)?
    
```{r}
m1 <- glm(exam_score ~ practice_hours,
          data = training)
```

- Note the lack of output from the above code -- we have *saved* the output to the object `m1`.

```{r}
m1 |> coefficients()
```

$$\hat{\text{exam}} = 57.25 + 3.11 \text{ practice}$$

## Interpreting Coefficients

- Interpretation of the slope, $\hat{\beta}$

    - For a [$k$] [units of $x$] increase in [$x$], we expect [$y$] to [increase or decrease] by [$k \times |\hat{\beta}_1|$] [units of $y$].

- Interpretation of the $y$-intercept, $\hat{\beta}_0$

    - When [$x$] is 0, we expect the mean of [$y$] to be [$\hat{\beta}_0$].

- Caveat:

    - It does not always make sense for $x=0$. In this situation, we do not interpret the $y$-intercept.
    
    - Some applications (e.g., psychology) will center the $x$ variable around its mean. Then, when $x=0$, we are interpreting in terms of the "average value of $x$."

## Example 1

- Recall our estimates,

```{r}
m1 |> coefficients()
```

- For a [$k$] [units of $x$] increase in [$x$], we expect [$y$] to [increase or decrease] by [$k \times |\hat{\beta}_1|$] [units of $y$].

    - For a 1 hour increase in practice time, we expect the exam score to increase by 3.11 points.
    
    - For a 3 hour increase in practice time, we expect the exam score to increase by 9.33
    
## Example 1

- Recall our estimates,

```{r}
m1 |> coefficients()
```    

- When [$x$] is 0, we expect the mean of [$y$] to be [$\hat{\beta}_0$].

    - When there is no practice, we expect the average exam score to be 57.25.
    
## Example 3

- A bakery owner in Savanna Central wants your help: "Some days, pastries practically fly off the shelves. Other days, we have trays left over. What actually helps explain our sales?"

- We have data from 40 days to answer the following question:

    - How are advertising spending ($x_1$), foot traffic ($x_2$), temperature ($x_3$), and discounts ($x_4$) related to daily bakery sales ($y$)?
    
```{r}
#| echo: false
bakery_sales |> head(n=3)
```

## Example 3   

- How are advertising spending ($x_1$), foot traffic ($x_2$), temperature ($x_3$), and discounts ($x_4$) related to daily bakery sales ($y$)?

```{r}
m3 <- glm(sales_dollars ~ advertising_dollars	+ foot_traffic + temperature_c + discount_percent,
          data = bakery_sales)
m3 |> coefficients()
```

$$\hat{\text{sales}} = 142.69 + 1.73 \text{ adv.} + 1.38 \text{ traff.} + 8.61 \text{ temp.} + 11.92 \text{ disc.}$$

## Example 3 

$$\hat{\text{sales}} = 142.69 + 1.73 \text{ adv.} + 1.38 \text{ traff.} + 8.61 \text{ temp.} + 11.92 \text{ disc.}$$

- Interpreting coefficients:

    - Before accounting for advertising expenditures, foot traffic, temperature, and discount offered, we expect bakery sales to be, on average, $142.69.

    - For a $1 increase in advertising, we expect bakery sales to increase by $1.73.
    
    - For a 1 citizen increase in foot traffic, we expect bakery sales to increase by $1.38.
    
    - For a 1^o^C increase in temperature, we expect bakery sales to increase by $8.61.
    
    - For a 1 percentage point increase in discount offered, we expect bakery sales to increase by $11.92.
    
## Wrap Up

- We have now seen the basics about correlation and regression analysis.

    - Correlation is unitless and bounded to [-1, 1].
    
        - This tells us the *strength* of the relationship.
    
    - Slopes have units and can take on any value on the real line.
    
        - This allows us to concretely quantify the increase or decrease.

- In the next lecture, we will look at making statistical inference on our regression models.