---
title: "General Linear Models:<br>Statistical Inference"
execute:
  echo: true
  warning: false
  message: false
  error: true
format:
  revealjs:
    theme: simple
    css: style_lecture.css
    include-after-body: headers.html
    self-contained: true
    slide-number: true
    center: false
    width: 1600
    height: 900
    df-print: kable
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
    code-annotations: hover
editor: source
---

## Introduction

- In our previous lectures, we discussed correlation and estimation of regression coefficients.

    - Correlation is unitless.
    
    - Regression coefficients retain the units of $x$ and $y$ for interpretations.

- In this lecture, we introduce statistical inference for regression:

    - Confidence intervals for $\beta_i$
    
    - Overall test for a significant regression line
    
    - Test for individual predictors

## `R` Set Up

```{r}
library(tidyverse)
library(ssstats)
library(broom)
training <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/7-1-training.csv")
wellness <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/7-2-wellness.csv")
reports <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/7-3-reports.csv")
break_room <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/7-4-break_room.csv")
bakery_sales <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/7-5-bakery_sales.csv")
cafe_sales <- read_csv("https://raw.githubusercontent.com/samanthaseals/SDSI/refs/heads/main/files/data/lectures/7-6-cafe_sales.csv")
```

## Regression Assumptions

- Recall the general linear model,

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + ... + \beta_k x_k + \varepsilon$$

- Like in ANOVA, we assume

$$\varepsilon \sim N(0, \sigma^2)$$

## Regression Assumptions (`R`)

- We will use the `reg_check()` function from `library(ssstats)`.

- Generic syntax:

```{r}
#| eval: false
model_name |> reg_check() 
```

## Example 1

- Recall the `training` dataset,

    - Is there a relationship between weekly practice hours ($x$) and exam score ($y$)?
    
```{r}
#| echo: false
head(training, n = 3)
```

## Example 1

- Is there a relationship between weekly practice hours ($x$) and exam score ($y$)?

```{r}
m1 <- glm(exam_score ~ practice_hours,
          data = training)
m1 |> coefficients()
```

## Example 1

- Checking assumptions,

```{r}
#| fig-align: 'center'
m1 |> reg_check()
```

## Example 2

- Recall the `reports` dataset,

    - Is there a relationship between officer height ($x$) and report accuracy ($y$)?

```{r}
#| echo: false
head(reports, n = 3)
```

## Example 2

- Is there a relationship between officer height ($x$) and report accuracy ($y$)?

- Constructing the model,

```{r}
m2 <- glm(report_accuracy ~ height_cm, 
          data = reports)
m2 |> coefficients()
```

## Example 2

- Checking assumptions,

```{r}
#| fig-align: 'center'
m2 |> reg_check()
```

## Statistical Inference in Linear Regression

- Once we confirm that our residuals follow a normal distribution, we begin to make inference.

    - Omnibus test for the model.
    
    - Individual predictor tests.
    
        - Confidence intervals for $\beta_i$.
        
- If the residuals do *not* follow a normal distribution with mean 0 and some common variance, $\sigma^2$, we do *not* follow our modeling up with formal hypothesis tests.        

## Confidence Intervals for $\beta_i$

- Like all statistics, we can put a confidence interval on $\beta_i$.

    - When reporting regression results, I *always* include the corresponding CI.
    
$$\hat{\beta}_i \pm t_{\alpha/2, n-k-1} \text{ SE}_{\hat{\beta}_i}$$



## Confidence Intervals for $\beta_i$ (`R`)

- We can use the `tidy()` function to get regression results with confidence intervals,

```{r}
#| eval: false
model_name |> tidy(conf.int = TRUE) 
```

## Example 1

- Returning to our first example,

    - Is there a relationship between weekly practice hours ($x$) and exam score ($y$)?
    
```{r}
#| eval: false
m1 |> tidy(conf.int = TRUE) 
```

```{r}
#| echo: false
m1 |> tidy(conf.int = TRUE) |> select(term, estimate, conf.low, conf.high)
```

## Example 3

- Recall the `bakery_sales` dataset,

    - How are advertising spending ($x_1$), foot traffic ($x_2$), temperature ($x_3$), and discounts ($x_4$) related to daily bakery sales ($y$)?
    
```{r}
#| echo: false
bakery_sales |> head(n=3)
```

## Example 3

- Recall the `bakery_sales` dataset,

    - How are advertising spending ($x_1$), foot traffic ($x_2$), temperature ($x_3$), and discounts ($x_4$) related to daily bakery sales ($y$)?
    
```{r}
#| eval: false
m3 |> tidy(conf.int = TRUE) 
```

```{r}
#| echo: false
m3 <- glm(sales_dollars ~ advertising_dollars   + foot_traffic + temperature_c + discount_percent,
          data = bakery_sales)
m3 |> tidy(conf.int = TRUE) |> select(term, estimate, conf.low, conf.high)
```

##  Hypothesis Testing: Significant Regression Line

- We can use the likelihood ratio test to test for a "significant regression" line. 

    - A significant regression line means that there is a non-zero slope among all slopes. 

- **Hypotheses**

    - $H_0: \ \beta_1 = ... = \beta_k = 0$ 
    - $H_1:$ at least one is different

- **Test Statistic and *p*-Value**

    - $\chi^2_0$ (from R)
    - $p = \text{P}[\chi^2_{k} \ge \chi^2_0]$
    
##  Hypothesis Testing: Significant Regression Line (R)

- We will use the `anova()` function to test for a significant model.

```{r}
#| eval: false
full <- glm(y ~ x_1 + x_2 + ... + x_k, data = dataset_name)
reduced <- glm(y ~ 1, data = dataset_name)
anova(reduced, full, test = "Chisq")
```

## Example 1

- Revisiting our model from Example 1, 

$$\hat{\text{exam}} = 57.25 + 3.11 \text{ practice}$$

```{r}
full <- glm(exam_score ~ practice_hours, data = training)
reduced <- glm(exam_score ~ 1, data = training)
anova(reduced, full, test = "Chisq")
```

## Example 1

- **Hypotheses:**

    - $H_0: \beta_1 = 0$ 
    - $H_1: \beta_1 \ne 0$
    
- **Test Statistic and $p$-Value:**

    - $\chi^2_0 = 2773.9$
    - $p < 0.001$
    
- **Rejection Region:**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$
    
- **Conclusion and Interpretation:**    

    - Reject $H_0$. There is sufficient evidence to suggest that the slope is non-zero.
    
## Example 3

- Revisiting our model from Example 3, 

$$\hat{\text{exam}} = 57.25 + 3.11 \text{ practice}$$

```{r}
full <- glm(sales_dollars ~ advertising_dollars	+ foot_traffic + temperature_c + discount_percent, data = bakery_sales)
reduced <- glm(sales_dollars ~ 1, data = bakery_sales)
anova(reduced, full, test = "Chisq")
```

## Example 3

- **Hypotheses:**

    - $H_0: \beta_1 = \beta_2 = \beta_3 = \beta_4 = 0$ 
    - $H_1:$ at least one $\beta_i \ne 0$
    
- **Test Statistic and $p$-Value:**

    - $\chi^2_0 = 712318$
    - $p < 0.001$
    
- **Rejection Region:**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$
    
- **Conclusion and Interpretation:**    

    - Reject $H_0$. There is sufficient evidence to suggest that at least one of the slopes is non-zero.


##  Hypothesis Testing: Individual Predictors

- We can use the $t$-test to test for "significant predictors"

    - A significant predictor means that there is a non-zero slope between the outcome and this specific $x_i$.
    
- **Hypotheses**

    - $H_0: \ \beta_i = 0$ 
    - $H_1: \ \beta_i \ne 0$

- **Test Statistic**

$$t_0 = \frac{\hat{\beta}_i}{\text{SE}_{\beta_i}} \sim t_1$$

##  Hypothesis Testing: Individual Predictors (`R`)

- We again turn to `tidy()` for output,

```{r}
#| eval: false
model_name |> tidy(conf.int = TRUE)
```

## Example 1

- Returning to our first example,

```{r}
m1 |> tidy(conf.int = TRUE)
```

- Practice time is a significant predictor of exam score ($p < 0.001$).


## Example 3

- Returning to our third example,

```{r}
m3 |> tidy(conf.int = TRUE)
```

- All predictors are significant (all $p < 0.001$).

## Example 4

- A cafe owner in Savanna Central brings you a spreadsheet: "We’ve been tracking everything! Can you help us figure out which of these are related to sales?"

- We have collected data from 40 days to answer the following question:

    - Which of the five explanatory variables are associated with daily cafe sales after accounting for the other variables?

```{r}
#| echo: false
cafe_sales |> head(n=3)
```

## Example 4

- Constructing the model,

    - Which of the five explanatory variables are associated with daily cafe sales after accounting for the other variables?

```{r}
m4 <- glm(sales_dollars ~ window_display_hours + music_volume_db + temperature_c + foot_traffic + advertising_dollars,
          data = cafe_sales)
m4 |> tidy(conf.int = TRUE)
```

## Example 4

- Is this a significant regression model?

```{r}
full <- glm(sales_dollars ~ window_display_hours + music_volume_db + temperature_c + foot_traffic + advertising_dollars, data = cafe_sales)
reduced <- glm(sales_dollars ~ 1, data = cafe_sales)
anova(reduced, full, test = "Chisq")
```


## Example 4

- **Hypotheses:**

    - $H_0: \beta_1 = \beta_2 = \beta_3 = \beta_4 = \beta_5 =0$ 
    - $H_1:$ at least one $\beta_i \ne 0$
    
- **Test Statistic and $p$-Value:**

    - $\chi^2_0 = 1245567$
    - $p < 0.001$
    
- **Rejection Region:**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$
    
- **Conclusion and Interpretation:**    

    - Reject $H_0$. There is sufficient evidence to suggest that at least one of the slopes is non-zero.

## Example 4

- Looking at individual predictors,

```{r}
m4 <- glm(sales_dollars ~ window_display_hours + music_volume_db + temperature_c + foot_traffic + advertising_dollars,
          data = cafe_sales)
m4 |> tidy(conf.int = TRUE)
```

## Example 4

- Looking at the individual predictors,

    - Window display is not a significant predictor ($p = 0.828$)
    
    - Music volume is not a significant predictor ($p = 0.925$)
    
    - Temperature is not a significant predictor ($p = 0.128$)
    
    - Foot traffic is a significant predictor ($p < 0.001$)
    
    - Advertising dollars is a significant predictor ($p < 0.001$)

## Wrap Up

- We have covered the basics of linear regression.

- In Statistics for Data Science II, you will learn more about statistical modeling.

    - A deeper dive into ordinary least squares.
    
        - Additional predictor types.
        
        - Model fit and diagnostics.
        
        - Data visualization.

    - Logistic regression.
    
    - Poisson and negative binomial regressions.